---
title: "Module 3: Model Evaluation & Selection"
subtitle: "Auditing Driver Acceptance for Unequal Error Rates"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.two-col { display: flex; align-items: flex-start; gap: 1em; }
.col-narrow { flex: 1; }
.col-wide   { flex: 2; }
.big-table table {
  font-size: 150%;
  width: 65%;
  margin: 0.8em auto;
}
.big-table table th, .big-table table td {
  padding: 0.5em 1em;
  text-align: center;
}
.cm-corner {
  position: absolute;
  top: 90px;
  right: 40px;
  border-collapse: collapse;
  font-size: 90%;
}
.cm-corner th, .cm-corner td {
  border: 1px solid #888;
  padding: 8px 14px;
  text-align: center;
  font-weight: bold;
}
.cm-tp { background: #c8e6c9; color: #1b5e20; }
.cm-fn { background: #ffe0b2; color: #e65100; }
.cm-fp { background: #ffcdd2; color: #b71c1c; }
.cm-tn { background: #bbdefb; color: #0d47a1; }
.tp { color: #1b5e20; font-weight: bold; }
.fn { color: #e65100; font-weight: bold; }
.fp { color: #b71c1c; font-weight: bold; }
.tn { color: #0d47a1; font-weight: bold; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
library(yardstick)
theme_set(theme_minimal(base_size = 14))
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Learning Problem</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">Linear Models</a></td><td>✓ done</td></tr>
<tr><td><b>3</b></td><td><b>Model Evaluation &amp; Selection</b> <i>(you are here)</i></td><td>← current</td></tr>
<tr><td>4</td><td>Tree-Based Methods</td><td>upcoming</td></tr>
<tr><td>6</td><td>Unsupervised Learning</td><td>upcoming</td></tr>
<tr><td>7</td><td>Neural Networks Foundations</td><td>upcoming</td></tr>
<tr><td>11</td><td>Fairness Frameworks &amp; Metrics</td><td>upcoming</td></tr>
<tr><td>12</td><td>Auditing &amp; Interpretability</td><td>upcoming</td></tr>
</table>

---

# One Number Is Not Enough

In Module 1 we used **MSE**. In Module 2 we used **accuracy**. Both are scalars — they hide *where* and *for whom* the model is wrong.

--

**Quick example**: if only 1% of events are positive, "always predict negative" gets **99% accuracy** — and tells you nothing.

We saw this hint in Module 2: every model had ~75% accuracy yet predicted very different acceptance rates per group. We need finer metrics to see *why*.

--

For classification, look at the **confusion matrix**:

.big-table[
|              | Predicted = 1 | Predicted = 0 |
|--------------|---------------|---------------|
| **Actual = 1**   | TP            | FN            |
| **Actual = 0**   | FP            | TN            |
]

---

# Metrics Derived from the Confusion Matrix

<table class="cm-corner">
<tr><th></th><th>Pred = 1</th><th>Pred = 0</th></tr>
<tr><th>Actual = 1</th><td class="cm-tp">TP</td><td class="cm-fn">FN</td></tr>
<tr><th>Actual = 0</th><td class="cm-fp">FP</td><td class="cm-tn">TN</td></tr>
</table>

- **Accuracy** = $\dfrac{\color{#1b5e20}{TP} + \color{#0d47a1}{TN}}{\color{#1b5e20}{TP} + \color{#e65100}{FN} + \color{#b71c1c}{FP} + \color{#0d47a1}{TN}}$ — overall fraction correct

--

- **Precision** = $\dfrac{\color{#1b5e20}{TP}}{\color{#1b5e20}{TP} + \color{#b71c1c}{FP}}$ — of what I predicted positive, how many were real?

--

- **True positive rate (recall)** = $\dfrac{\color{#1b5e20}{TP}}{\color{#1b5e20}{TP} + \color{#e65100}{FN}}$ — of the real positives, how many did I predict positive?

--

- **False positive rate** = $\dfrac{\color{#b71c1c}{FP}}{\color{#b71c1c}{FP} + \color{#0d47a1}{TN}}$ — of the real negatives, how many did I predict positive?

--

- **F1** = $2 \cdot \dfrac{\text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}}$ — harmonic mean (penalizes lopsided models)

--

**Worked example.** Driver-acceptance model on 1000 ride requests (positive = "accepted"): TP = 480, FN = 220, FP = 130, TN = 170.

$$\text{accuracy} = \tfrac{480 + 170}{1000} = 0.65 \quad
\text{precision} = \tfrac{480}{480 + 130} \approx 0.79 \quad
\text{recall} = \tfrac{480}{480 + 220} \approx 0.69 \quad
\text{FPR} = \tfrac{130}{300} \approx 0.43$$

---

# Which Metric When?

*"Classes" = the possible values of $Y$ (e.g., accepted vs not accepted).*

- **Imbalanced classes** → recall + precision
  - At 4 AM only 5% of requests get accepted. "Always predict reject" → accuracy = 0.95, recall = 0, precision undefined (0/0). Useless.
- **The probability value itself matters** → log loss
  - Dispatch sends a driver only if $P(\text{accept}) > 0.6$. If the model says 0.7 but the true rate is 0.4, the rule misfires — even if the model gets the *ordering* of requests right.
- **Cost-sensitive** → weighted loss
  - Predicting "accept" when the driver actually rejects → wasted dispatch (~$1). Predicting "reject" when they would have accepted → lost ride (~$5). False negatives cost **5× more**.

---

# ROC Curves and AUC

**ROC** = *Receiver Operating Characteristic* (from WWII radar). **AUC** = *Area Under the Curve*.

A classifier outputs a probability. Pick a **threshold** to turn it into 0/1; sweeping the threshold traces the ROC curve. AUC = area under it.

- **AUC = 0.5** → random; **AUC = 1.0** → perfect ranking
- Equivalent reading: probability that a random positive scores higher than a random negative

```{r roc-demo, echo=FALSE, fig.height=2.8}
set.seed(7)
make_roc <- function(separation) {
  pos <- rnorm(500, mean = separation, sd = 1)
  neg <- rnorm(500, mean = 0, sd = 1)
  scores <- c(pos, neg)
  truth  <- factor(c(rep(1, 500), rep(0, 500)), levels = c(0, 1))
  yardstick::roc_curve(tibble(truth = truth, score = scores),
                       truth, score, event_level = "second") |>
    mutate(model = sprintf("AUC ≈ %.2f",
      yardstick::roc_auc_vec(truth, scores, event_level = "second")))
}

bind_rows(make_roc(0), make_roc(1), make_roc(2.5)) |>
  ggplot(aes(1 - specificity, sensitivity, color = model)) +
  geom_line(linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  labs(title = "Three classifiers at different AUC values",
       x = "False Positive Rate", y = "True Positive Rate", color = NULL) +
  theme(legend.position = "right")
```

---

# Same Story, with Threshold on the X-Axis

```{r threshold-view, echo=FALSE, fig.height=3.4}
set.seed(7)
build_curves <- function(separation, label) {
  # Raw scores then squashed through sigmoid → probabilities in [0, 1]
  pos <- plogis(rnorm(500, mean = separation, sd = 1))
  neg <- plogis(rnorm(500, mean = 0, sd = 1))
  thr <- seq(0, 1, length.out = 200)
  tibble(
    threshold = rep(thr, 2),
    metric    = rep(c("TPR", "FPR"), each = length(thr)),
    value     = c(sapply(thr, \(t) mean(pos >= t)),
                  sapply(thr, \(t) mean(neg >= t))),
    model     = label
  )
}

bind_rows(
  build_curves(0,   "AUC ≈ 0.53"),
  build_curves(1,   "AUC ≈ 0.74"),
  build_curves(2.5, "AUC ≈ 0.96")
) |>
  ggplot(aes(threshold, value, color = metric)) +
  geom_line(linewidth = 1.1) +
  facet_wrap(~ model) +
  scale_color_manual(values = c("TPR" = "darkgreen", "FPR" = "firebrick")) +
  labs(x = "Decision threshold (probability)", y = "Rate", color = NULL,
       title = "TPR and FPR as the threshold sweeps")
```

For each classifier, **two curves** of TPR (green) and FPR (red) vs threshold. The ROC curve from the previous slide is built by plotting these two curves *against each other*, eliminating the threshold from the picture.

---

# k-Fold Cross-Validation

CV doesn't estimate model **parameters** ($\beta$ etc. — those are fit inside each fold). It estimates the model's **generalization error** by averaging over many train/validation splits.

--

1. Split training data into $k$ folds
2. For each fold: train on the other $k-1$, predict on the held-out one
3. Average the metric across folds

--

Then use the CV error to pick a **hyperparameter** (like Ridge's $\lambda$, polynomial degree, tree depth):

```
for each candidate λ:
    for each of the k folds:
        fit model with this λ on the other k-1 folds
        record error on held-out fold
    average the k errors
pick the λ with the lowest average CV error
```

- $k = 5$ or $k = 10$ are standard
- Module 1 used a *single* validation split for polynomial degree; CV is the more robust version

---

# A Perfect AUC Can Still Be Useless

Consider a model that outputs these scores on 5 ride requests (actual label in parens):

| score | label    |
|-------|----------|
| 0.21  | accepted |
| 0.20  | accepted |
| 0.19  | rejected |
| 0.18  | accepted |
| 0.17  | rejected |

--

Every accepted request scored higher than every rejected one → **AUC = 1.0** (perfect ranking).

--

But the predicted probabilities sit around **0.2**, even though the actual acceptance rate is 60%. If your dispatch rule is "send a driver when $P(\text{accept}) > 0.5$", **you'd dispatch zero drivers** — despite the model perfectly knowing which requests would be accepted.

--

AUC only cares about **order**. It doesn't care that the *numbers are wrong*. We need a different concept for that → **calibration**.

---

# Calibration

A classifier is **well-calibrated** if, **for every** $x \in [0, 1]$, among ride requests where the model predicted $P(\text{accept}) \approx x$, the fraction actually accepted is also $\approx x$.

$$P(Y = 1 \mid \hat{p}(X) = x) = x \quad \text{for all } x$$

--

**Worked example at $x = 0.8$.** Take all ride requests where the model predicted $\hat{p} \in [0.75, 0.85]$ — say there are 200 such requests. Count the actual acceptances:

- ≈ 160 accepted → ratio ≈ 0.80 → **well-calibrated** ✓
- only ≈ 100 accepted → ratio ≈ 0.50 → **overconfident** (model says 0.8, reality is 0.5)
- ≈ 190 accepted → ratio ≈ 0.95 → **underconfident**

For the model to be calibrated *overall*, the same property must hold at every $x$ — that's what the reliability diagram below checks.

--

- Check it with a **reliability diagram**: bin predicted probabilities, plot bin mean vs observed rate. Diagonal = perfect.
- Fix it with **Platt scaling** (logistic regression on the scores) or **isotonic regression** after the fact.

---

# Reliability Diagrams: Three Examples

<div class="two-col">
<div class="col-narrow">
<ul>
<li><b style="color:darkgreen">Green</b> lies on the diagonal → predicted = observed at every <i>x</i> → <b>calibrated</b></li>
<li><b style="color:firebrick">Red</b> is <i>below</i> the diagonal → predicts higher than reality → <b>overconfident</b></li>
<li><b style="color:steelblue">Blue</b> is <i>above</i> the diagonal → predicts lower than reality → <b>underconfident</b></li>
</ul>
</div>
<div class="col-wide">

```{r reliability-diagrams, echo=FALSE, fig.height=4.2, fig.width=6}
set.seed(11)
n_sim <- 5000

# Three classifiers with the same AUC but different calibration
true_p <- runif(n_sim, 0, 1)
y <- rbinom(n_sim, 1, true_p)

models <- tibble(
  true_p = true_p,
  y = y,
  `Well-calibrated`   = true_p,                          # identity
  Overconfident       = plogis(2 * qlogis(true_p)),      # pushes toward 0/1
  Underconfident      = plogis(0.5 * qlogis(true_p))     # pulls toward 0.5
) |>
  pivot_longer(`Well-calibrated`:Underconfident,
               names_to = "model", values_to = "p_hat")

bins <- models |>
  mutate(bin = cut(p_hat, breaks = seq(0, 1, by = 0.1),
                   include.lowest = TRUE)) |>
  group_by(model, bin) |>
  summarise(predicted = mean(p_hat),
            observed  = mean(y),
            n = n(), .groups = "drop")

ggplot(bins, aes(predicted, observed, color = model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Well-calibrated" = "darkgreen",
                                "Overconfident"   = "firebrick",
                                "Underconfident"  = "steelblue")) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Predicted P(accept)", y = "Actual acceptance rate", color = NULL,
       title = NULL)
```

</div>
</div>

---

# Fixing Calibration After the Fact

Both methods are **post-fit recipes**: don't retrain the original model — train a tiny second model that maps `old score → calibrated probability`, using a held-out calibration set.

--

**Platt scaling** — assume the distortion is a smooth sigmoid:

1. Take held-out scores $s$ and labels $y$
2. Fit a 1-D logistic regression $y \sim s$ → get $a, b$
3. Replace each score with $\hat p_{\text{new}} = \sigma(a \cdot s + b)$

--

**Isotonic regression** — let the data choose the shape, only require it to be monotonic:

1. Same held-out $(s, y)$
2. Fit a monotonic step function $g(s)$ to $y$
3. Replace each score with $\hat p_{\text{new}} = g(s)$

--

| | Platt | Isotonic |
|---|---|---|
| Shape | Sigmoid (parametric) | Any monotonic (non-parametric) |
| Data needed | A few hundred points | A few thousand points |
| Risk | Underfit if distortion isn't sigmoid | Overfit when data is small |

R: `probably::cal_estimate_logistic()` / `probably::cal_estimate_isotonic()`

---
class: inverse, center, middle

# Discrimination Angle
### Equal AUC ≠ Equal Treatment

---

# Unequal Error Rates

A classifier can have **great overall accuracy** and still be terrible for a specific group. Two distinct ways:

--

1. **Different precision** — the same flag means different things: a "suspicious" label has 90% precision for one group and 50% for another

--

2. **Different recall** (or FPR) — the model catches positives for one group but misses them for another, or flags innocent people more in one group

--

3. **Different calibration** — a "score of 0.8" actually means 60% risk for group A and 90% for group B

---

# The COMPAS Story (2016)

ProPublica audited COMPAS, a recidivism risk-scoring tool used in U.S. courts.

--

- Overall AUC was **similar** for Black and white defendants
- Among defendants who did **not** re-offend, Black defendants were classified high-risk **twice as often** (higher FPR)
- Among defendants who **did** re-offend, white defendants were classified low-risk more often (lower TPR for whites)

--

The COMPAS makers responded: the model was **calibrated equally** across groups (a score of 7 meant the same recidivism rate for both).

--

**Both sides were right.** This is the *impossibility result* we'll see in Module 11: you can't satisfy all fairness criteria simultaneously.

---

# What "Auditing" Actually Means

A real audit of an ML system is just: **compute every metric you care about, broken down by demographic group.**

--

You don't need new statistics — you need the discipline to slice the existing ones.

--

**Example.** A driver-acceptance model with overall accuracy 0.74 and AUC 0.81. The audit reveals:

| Group         | Accuracy | Precision | Recall | FPR  |
|---------------|----------|-----------|--------|------|
| Non-minority  | 0.78     | 0.86      | 0.85   | 0.34 |
| Minority      | 0.69     | 0.61      | 0.62   | 0.21 |

Overall AUC looks fine, but the model **predicts "accept" for minority requests much less often** — recall drops from 0.85 to 0.62, and precision from 0.86 to 0.61. Same model, very different per-group experience.

--

In R: `group_by(group) |> yardstick::metrics(truth, prediction)`. Done.

---
class: inverse, center, middle

# Exercise
### Auditing the Driver-Acceptance Model

---

# Part 1: Reuse the Module 2 Setup

```{r data-gen}
set.seed(2024); n <- 5000
neighborhoods <- tibble(
  neighborhood   = c("Downtown","Midtown","Eastside","Southside"),
  base_accept    = c(0.85, 0.78, 0.55, 0.40),
  pct_minority   = c(0.20, 0.35, 0.70, 0.85)
)
requests <- tibble(
  request_id = 1:n,
  neighborhood = sample(neighborhoods$neighborhood, n, TRUE,
                        prob = c(0.35, 0.30, 0.20, 0.15))
) |>
  left_join(neighborhoods, by = "neighborhood") |>
  mutate(
    hour = sample(0:23, n, replace = TRUE),
    is_night = as.integer(hour >= 22 | hour <= 5),
    trip_distance_mi = pmax(rlnorm(n, 1.2, 0.6), 0.3),
    rider_rating = pmin(pmax(rnorm(n, 4.7, 0.4), 1), 5)
  )
```

---

# Part 1: Generate Acceptance + Demographics

```{r data-gen-2}
requests <- requests |>
  mutate(
    logit_p = qlogis(base_accept) - 0.6 * is_night -
              0.15 * abs(trip_distance_mi - 4) +
              0.5 * (rider_rating - 4.7),
    accepted = rbinom(n, 1, plogis(logit_p)),
    is_minority = rbinom(n, 1, pct_minority)
  )
```

```{r baseline, echo=FALSE}
requests |>
  group_by(is_minority) |>
  summarise(n = n(),
            accept_rate = round(mean(accepted), 3),
            .groups = "drop") |>
  knitr::kable()
```

---

# Part 2: Fit the Race-Blind Classifier

```{r fit}
set.seed(123)
train_idx <- sample(1:n, 0.7 * n)
train <- requests[train_idx, ]; test <- requests[-train_idx, ]

# is_minority is NOT in the formula (same as Module 2)
fit <- glm(
  accepted ~ neighborhood + is_night + trip_distance_mi + rider_rating,
  data = train, family = binomial
)

test$pred_prob  <- predict(fit, test, type = "response")
test$pred_class <- factor(as.integer(test$pred_prob >= 0.5), levels = c(0, 1))
test$truth      <- factor(test$accepted, levels = c(0, 1))
```

---

# Part 3: Overall Metrics

.small[
```{r overall, echo=FALSE}
bind_rows(
  accuracy(test, truth, pred_class),
  precision(test, truth, pred_class, event_level = "second"),
  recall(test, truth, pred_class, event_level = "second"),
  f_meas(test, truth, pred_class, event_level = "second"),
  roc_auc(test, truth, pred_prob, event_level = "second")
) |>
  mutate(.estimate = round(.estimate, 3)) |>
  knitr::kable()
```
]

--

Looks fine. Now disaggregate.

---

# Part 4: Per-Group Metrics

.small[
```{r per-group, echo=FALSE}
test |>
  group_by(is_minority) |>
  summarise(
    n         = n(),
    accuracy  = round(accuracy_vec(truth, pred_class), 3),
    precision = round(precision_vec(truth, pred_class, event_level = "second"), 3),
    recall    = round(recall_vec(truth, pred_class, event_level = "second"), 3),
    fpr       = {
      tn <- sum(truth == 0 & pred_class == 0)
      fp <- sum(truth == 0 & pred_class == 1)
      round(fp / (fp + tn), 3)
    },
    auc       = round(roc_auc_vec(truth, pred_prob, event_level = "second"), 3),
    .groups   = "drop"
  ) |>
  knitr::kable()
```
]

--

AUC is similar across groups, but the **false positive rate** and **flag rate** are very different.

---

# Per-Group ROC Curves

```{r roc-plot, echo=FALSE, fig.height=4}
test |>
  group_by(is_minority) |>
  roc_curve(truth, pred_prob, event_level = "second") |>
  mutate(group = ifelse(is_minority == 1, "Minority", "Non-minority")) |>
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = group)) +
  geom_line(linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  scale_color_manual(values = c("Non-minority" = "steelblue",
                                "Minority" = "firebrick")) +
  labs(title = "Per-Group ROC Curves",
       x = "False Positive Rate", y = "True Positive Rate", color = "Group")
```

---

# Per-Group Calibration

```{r calib-plot, echo=FALSE, fig.height=4}
test |>
  mutate(bin = cut(pred_prob, breaks = seq(0, 1, by = 0.1),
                   include.lowest = TRUE)) |>
  group_by(is_minority, bin) |>
  summarise(avg_predicted = mean(pred_prob),
            actual_rate = mean(accepted), n = n(), .groups = "drop") |>
  mutate(group = ifelse(is_minority == 1, "Minority", "Non-minority")) |>
  ggplot(aes(x = avg_predicted, y = actual_rate, color = group)) +
  geom_line(linewidth = 1) + geom_point(aes(size = n)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  scale_color_manual(values = c("Non-minority" = "steelblue",
                                "Minority" = "firebrick")) +
  labs(title = "Per-Group Calibration",
       x = "Predicted Acceptance Probability", y = "Actual Acceptance Rate",
       color = "Group", size = "n")
```

---

# Disparate Impact at Every Threshold

```{r threshold-sweep, echo=FALSE, fig.height=3.8}
thresholds <- seq(0.05, 0.95, by = 0.05)
threshold_audit <- map_dfr(thresholds, function(t) {
  test |>
    mutate(flag = as.integer(pred_prob >= t)) |>
    group_by(is_minority) |>
    summarise(flag_rate = mean(flag), .groups = "drop") |>
    pivot_wider(names_from = is_minority, values_from = flag_rate,
                names_prefix = "g") |>
    mutate(threshold = t, di_ratio = g1 / g0)
})

ggplot(threshold_audit, aes(x = threshold, y = di_ratio)) +
  geom_line(linewidth = 1.2, color = "firebrick") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey") +
  labs(title = "Disparate Impact Ratio Across Thresholds",
       subtitle = "Picking a threshold is itself a fairness decision",
       x = "Decision Threshold", y = "DI Ratio (1.0 = parity)")
```

---
class: inverse

# The Key Questions

<br>

### 1. If two models have the same AUC, are they equally fair?

--

<br>

### 2. Should we pick the threshold that maximizes accuracy, or the one that minimizes the FPR gap?

--

<br>

### 3. What does "calibration" buy us if it can coexist with unequal FPRs?

---

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Learning Problem</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">Linear Models</a></td><td>✓ done</td></tr>
<tr><td>3</td><td>Model Evaluation &amp; Selection <i>(just finished)</i></td><td>✓ done</td></tr>
<tr><td><b>4</b></td><td><b>Tree-Based Methods</b></td><td>next</td></tr>
<tr><td>6</td><td>Unsupervised Learning</td><td>upcoming</td></tr>
<tr><td>7</td><td>Neural Networks Foundations</td><td>upcoming</td></tr>
<tr><td>11</td><td>Fairness Frameworks &amp; Metrics</td><td>upcoming</td></tr>
<tr><td>12</td><td>Auditing &amp; Interpretability</td><td>upcoming</td></tr>
</table>

Say **"start module 4"** when ready.
