---
title: "Module 1: The Learning Problem"
subtitle: "Disparate Wait Times in Ride-Sharing"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small .remark-code, .small table { font-size: 70%; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.5, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
theme_set(theme_minimal(base_size = 14))
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td><b>1</b></td><td><b>The Learning Problem</b> <i>(you are here)</i></td><td>← current</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">Linear Models</a></td><td>✓</td></tr>
<tr><td>3</td><td><a href="../module-03/slides.html">Model Evaluation &amp; Selection</a></td><td>✓</td></tr>
<tr><td>4</td><td>Tree-Based Methods</td><td>upcoming</td></tr>
<tr><td>6</td><td>Unsupervised Learning</td><td>upcoming</td></tr>
<tr><td>7</td><td>Neural Networks Foundations</td><td>upcoming</td></tr>
<tr><td>11</td><td><a href="../module-11/slides.html">Fairness Frameworks &amp; Metrics</a></td><td>✓ done</td></tr>
<tr><td>12</td><td>Auditing &amp; Interpretability</td><td>upcoming</td></tr>
</table>

---

# The Setup

We observe data $(X, Y)$ and want to learn a function $f$ such that

$$Y = f(X) + \varepsilon, \quad \mathbb{E}[\varepsilon] = 0, \quad \text{Var}(\varepsilon) = \sigma^2$$

--

- Choose $\hat{f}$ from some model class by minimizing a **loss function** on training data

--

- Evaluate on **held-out** data (validation / test)

--

- $\varepsilon$ is irreducible — no model can eliminate it

---

# Loss Functions

| Task | Loss | Formula |
|------|------|---------|
| Regression | MSE | $\frac{1}{n} \sum_i (y_i - \hat{y}_i)^2$ |
| Classification | 0-1 loss | $\frac{1}{n} \sum_i \mathbb{I}(y_i \neq \hat{y}_i)$ |
| Classification | Cross-entropy | $-\sum_i y_i \log(\hat{p}_i)$ |

--

- MSE penalizes large errors more heavily

--

- Cross-entropy rewards **well-calibrated probabilities** — preferred for probabilistic classifiers

---

# Train / Validation / Test

| Set | Purpose | When used |
|-----|---------|-----------|
| **Training** | Fit the model parameters | During training |
| **Validation** | Choose hyperparameters, compare models | During model selection |
| **Test** | Final unbiased performance estimate | Once, at the end |

--

Cross-validation approximates the validation step by rotating through folds.

---

# Example: Validation Picks the Hyperparameter

Train three models on the **same** training set, evaluate on **validation**:

```{r validation-example, echo=TRUE, results='hide'}
set.seed(99)
demo <- tibble(x = runif(200, 0, 10),
               y = sin(x) + rnorm(200, 0, 0.3))
demo_train <- demo[1:140, ]
demo_val   <- demo[141:200, ]

# Model A: degree-5 polynomial (less flexible)
fit_A <- lm(y ~ poly(x, 5), data = demo_train)

# Model B: degree-8 polynomial (moderate flexibility)
fit_B <- lm(y ~ poly(x, 8), data = demo_train)

# Model C: degree-20 polynomial (very flexible)
fit_C <- lm(y ~ poly(x, 20), data = demo_train)
```

---

# Validation Results

```{r validation-results, echo=FALSE}
# Monte Carlo: simulate 500 training sets to estimate Bias² and Variance
# Use only interior validation points (x in [2,8]) to avoid polynomial edge effects
n_sim <- 500
interior <- which(demo_val$x >= 2 & demo_val$x <= 8)
x_val <- demo_val$x[interior]
f_true <- sin(x_val)

preds_A <- preds_B <- preds_C <- matrix(NA, nrow = n_sim, ncol = length(x_val))
for (s in 1:n_sim) {
  x_sim <- runif(140, 0, 10)
  sim <- tibble(x = x_sim, y = sin(x) + rnorm(140, 0, 0.3))
  preds_A[s, ] <- predict(lm(y ~ poly(x, 5), data = sim), demo_val[interior, ])
  preds_B[s, ] <- predict(lm(y ~ poly(x, 8), data = sim), demo_val[interior, ])
  preds_C[s, ] <- predict(lm(y ~ poly(x, 20), data = sim), demo_val[interior, ])
}

bv <- function(preds, f_true) {
  mean_pred <- colMeans(preds)
  bias2 <- mean((f_true - mean_pred)^2)
  variance <- mean(apply(preds, 2, var))
  c(bias2, variance)
}

bv_A <- bv(preds_A, f_true)
bv_B <- bv(preds_B, f_true)
bv_C <- bv(preds_C, f_true)

tibble(
  Model = c("A (degree 5)", "B (degree 8)", "C (degree 20)"),
  `# Params` = c(6, 9, 21),
  `Train MSE`  = round(c(mean((demo_train$y - predict(fit_A, demo_train))^2),
                          mean((demo_train$y - predict(fit_B, demo_train))^2),
                          mean((demo_train$y - predict(fit_C, demo_train))^2)), 4),
  `Val MSE`    = round(c(mean((demo_val$y - predict(fit_A, demo_val))^2),
                          mean((demo_val$y - predict(fit_B, demo_val))^2),
                          mean((demo_val$y - predict(fit_C, demo_val))^2)), 4)
) |> knitr::kable()
```

--

- Model C wins on **training** data (lowest Train MSE) — it memorizes best

--

- Model B wins on **validation** data (lowest Val MSE) — it generalizes best

--

- Too simple (A) → underfits. Too complex (C) → overfits. **Validation finds the sweet spot (B).**

---

# Example: What Validation Reveals

```{r validation-plot, echo=FALSE, fig.height=4.8}
x_grid <- tibble(x = seq(0, 10, length.out = 300))
x_grid$pred_A <- predict(fit_A, x_grid)
x_grid$pred_B <- predict(fit_B, x_grid)
x_grid$pred_C <- predict(fit_C, x_grid)

# Annotation point: validation point i=15 (x~9.17)
i_pt <- 18
x0 <- demo_val$x[i_pt]
y0 <- demo_val$y[i_pt]
f0 <- sin(x0)
pred_A0 <- predict(fit_A, tibble(x = x0))

# Base plot (reused for zoom)
p_base <- ggplot() +
  geom_point(data = demo_train, aes(x, y, shape = "Training"), alpha = 0.3, size = 1) +
  geom_point(data = demo_val, aes(x, y, shape = "Validation"), color = "darkgreen", size = 2) +
  scale_shape_manual(values = c("Training" = 16, "Validation" = 17)) +
  geom_line(data = x_grid, aes(x, pred_A, color = "A: degree 5"), linewidth = 1) +
  geom_line(data = x_grid, aes(x, pred_B, color = "B: degree 8"), linewidth = 1) +
  geom_line(data = x_grid, aes(x, pred_C, color = "C: degree 20"), linewidth = 1) +
  stat_function(fun = sin, aes(color = "True f(x) = sin(x)"),
                linetype = "dashed", linewidth = 0.8)

p_base +
  labs(title = "Validation selects the hyperparameter (polynomial degree)",
       subtitle = "Next slide zooms in at the dotted line to show bias and variance",
       x = "x", y = "y", color = "Fit", shape = "Data") +
  theme(legend.position = "top")
```

---
class: inverse, center, middle

# Bias-Variance Tradeoff

---

# The Decomposition

For any estimate $\hat{f}$, the expected prediction error at a point $x_0$ is:

$$\mathbb{E}\left[(Y - \hat{f}(x_0))^2\right] = \underbrace{\left[\text{Bias}(\hat{f}(x_0))\right]^2}_{\text{systematic error}} + \underbrace{\text{Var}(\hat{f}(x_0))}_{\text{sensitivity to training set}} + \underbrace{\sigma^2}_{\text{irreducible}}$$

--

where:

- $\text{Bias}(\hat{f}(x_0)) = f(x_0) - \mathbb{E}[\hat{f}(x_0)]$

- $\text{Var}(\hat{f}(x_0)) = \mathbb{E}\left[(\hat{f}(x_0) - \mathbb{E}[\hat{f}(x_0)])^2\right]$

---

# Proof — Step 1: Separate Noise from Estimation

Start by substituting $Y = f(x_0) + \varepsilon$:

$$\mathbb{E}[(Y - \hat{f})^2] = \mathbb{E}[((f - \hat{f}) + \varepsilon)^2]$$

--

Expand the square:

$$= \mathbb{E}[(f - \hat{f})^2] + 2\,\mathbb{E}[(f - \hat{f})\,\varepsilon] + \mathbb{E}[\varepsilon^2]$$

--

The cross term vanishes: $\varepsilon$ is independent of $\hat{f}$ and $\mathbb{E}[\varepsilon] = 0$

$$= \mathbb{E}[(f(x_0) - \hat{f}(x_0))^2] + \sigma^2$$

---

# Proof — Step 2: Add and Subtract the Mean

The trick: insert $\mathbb{E}[\hat{f}]$ into the estimation error:

$$f - \hat{f} = \underbrace{(f - \mathbb{E}[\hat{f}])}_{\text{Bias}} + \underbrace{(\mathbb{E}[\hat{f}] - \hat{f})}_{\text{deviation from mean}}$$

--

Square and take expectation:

$$\mathbb{E}[(f - \hat{f})^2] = \text{Bias}^2 + 2 \cdot \text{Bias} \cdot \underbrace{\mathbb{E}[\mathbb{E}[\hat{f}] - \hat{f}]}_{= 0} + \underbrace{\mathbb{E}[(\hat{f} - \mathbb{E}[\hat{f}])^2]}_{\text{Var}(\hat{f})}$$

--

The cross term vanishes because $\mathbb{E}[\hat{f} - \mathbb{E}[\hat{f}]] = 0$.

---

# Proof — Result

$$\boxed{\mathbb{E}\left[(Y - \hat{f}(x_0))^2\right] = \left[\text{Bias}(\hat{f}(x_0))\right]^2 + \text{Var}(\hat{f}(x_0)) + \sigma^2}$$

--

| Component | Simple model (e.g., linear) | Complex model (e.g., degree-20 poly) |
|-----------|---------------------------|--------------------------------------|
| Bias² | High | Low |
| Variance | Low | High |

--

The **sweet spot** minimizes their sum → this is model selection.

---

# Back to Our Models — Where Are We Zooming In?

```{r zoom-context-plain, echo=FALSE, fig.height=4.8}
zoom_ylim <- c(-1.1, -0.37)

p_base +
  labs(title = "Validation selects the hyperparameter (polynomial degree)",
       x = "x", y = "y", color = "Fit", shape = "Data") +
  theme(legend.position = "top")
```

---
count: false

# Back to Our Models — Where Are We Zooming In?

```{r zoom-context-box, echo=FALSE, fig.height=4.8}
p_base +
  annotate("rect", xmin = x0 - 1.2, xmax = x0 + 1.2,
           ymin = zoom_ylim[1], ymax = zoom_ylim[2],
           fill = NA, color = "black", linewidth = 0.6, linetype = "dashed") +
  labs(title = "Validation selects the hyperparameter (polynomial degree)",
       x = "x", y = "y", color = "Fit", shape = "Data") +
  theme(legend.position = "top")
```

---

# Seeing It: 100 Simulated Training Sets

```{r validation-zoom, echo=FALSE, fig.height=4.2}
# Simulate 100 training sets → predictions at x0 for ALL three models (same training data)
set.seed(42)
n_sim <- 100
sim_preds <- sim_preds_B <- sim_preds_C <- numeric(n_sim)
for (s in 1:n_sim) {
  x_sim <- runif(140, 0, 10)
  sim_data <- tibble(x = x_sim, y = sin(x) + rnorm(140, 0, 0.3))
  sim_preds[s]   <- predict(lm(y ~ poly(x, 5), data = sim_data), tibble(x = x0))
  sim_preds_B[s] <- predict(lm(y ~ poly(x, 8), data = sim_data), tibble(x = x0))
  sim_preds_C[s] <- predict(lm(y ~ poly(x, 20), data = sim_data), tibble(x = x0))
}
ef0 <- mean(sim_preds); sd0 <- sd(sim_preds)
ef0_B <- mean(sim_preds_B); sd0_B <- sd(sim_preds_B)
ef0_C <- mean(sim_preds_C); sd0_C <- sd(sim_preds_C)

# Shared y-limits across all three slides
shared_ylim <- c(min(f0, y0, sim_preds, sim_preds_B, sim_preds_C) - 0.1,
                 max(f0, y0, sim_preds, sim_preds_B, sim_preds_C) + 0.1)

# Jitter the 100 predictions horizontally around x0
sim_df <- tibble(x = x0 + runif(n_sim, -0.15, 0.15), y = sim_preds)

# Simpler base: only true f and Model A
# Consistent color mapping across all zoomed plots
model_colors <- scales::hue_pal()(4)
names(model_colors) <- c("A: degree 5", "B: degree 8", "C: degree 20", "True f(x) = sin(x)")

p_simple <- ggplot() +
  geom_point(data = demo_train, aes(x, y, shape = "Training"), alpha = 0.3, size = 1.5) +
  geom_point(data = demo_val, aes(x, y, shape = "Validation"), color = "darkgreen", size = 2) +
  scale_shape_manual(values = c("Training" = 16, "Validation" = 17)) +
  geom_line(data = x_grid, aes(x, pred_A, color = "A: degree 5"), linewidth = 1) +
  stat_function(fun = sin, aes(color = "True f(x) = sin(x)"),
                linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(values = model_colors)

p_simple +
  # Show 100 predictions as orange dots
  geom_point(data = sim_df, aes(x = x, y = y), color = "orange", alpha = 0.5, size = 2.5) +
  # Highlight our actual training set's prediction in red (it's on the red line)
  geom_point(aes(x = x0, y = pred_A0), size = 4, color = "red", shape = 16) +
  # E[f-hat(x)] — mean of the 100 predictions
  geom_point(aes(x = x0, y = ef0), size = 6, shape = 4, color = "orange", stroke = 2) +
  coord_cartesian(xlim = c(x0 - 1.2, x0 + 1.2),
                  ylim = shared_ylim) +
  # Bias bracket: f(x) to E[f-hat(x)]
  annotate("segment", x = x0 + 0.3, xend = x0 + 0.3, y = f0, yend = ef0,
           color = "firebrick", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.22, xend = x0 + 0.38, y = f0, yend = f0,
           color = "firebrick", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.22, xend = x0 + 0.38, y = ef0, yend = ef0,
           color = "firebrick", linewidth = 0.8) +
  # Epsilon bracket: f(x) to y
  annotate("segment", x = x0 - 0.3, xend = x0 - 0.3, y = f0, yend = y0,
           color = "darkgreen", linewidth = 0.8) +
  annotate("segment", x = x0 - 0.38, xend = x0 - 0.22, y = f0, yend = f0,
           color = "darkgreen", linewidth = 0.8) +
  annotate("segment", x = x0 - 0.38, xend = x0 - 0.22, y = y0, yend = y0,
           color = "darkgreen", linewidth = 0.8) +
  # Variance bracket: ±1 SD around E[f-hat(x)]
  annotate("segment", x = x0 + 0.6, xend = x0 + 0.6,
           y = ef0 - sd0, yend = ef0 + sd0,
           color = "steelblue", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.52, xend = x0 + 0.68,
           y = ef0 - sd0, yend = ef0 - sd0,
           color = "steelblue", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.52, xend = x0 + 0.68,
           y = ef0 + sd0, yend = ef0 + sd0,
           color = "steelblue", linewidth = 0.8) +
  # Color-coded legend in bottom-right
  annotate("rect", xmin = x0 + 0.72, xmax = x0 + 1.18,
           ymin = shared_ylim[1] + 0.01, ymax = shared_ylim[1] + 0.40,
           fill = "white", color = "grey40", linewidth = 0.4) +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.33, hjust = 0,
           label = paste0("E[f-hat(x)] = ", round(ef0, 2)),
           size = 3.5, color = "orange", fontface = "bold") +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.25, hjust = 0,
           label = paste0("Bias = ", round(ef0 - f0, 2)),
           size = 3.5, color = "firebrick", fontface = "bold") +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.17, hjust = 0,
           label = paste0("SD = ", round(sd0, 3)),
           size = 3.5, color = "steelblue", fontface = "bold") +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.09, hjust = 0,
           label = paste0("epsilon = ", round(y0 - f0, 2)),
           size = 3.5, color = "darkgreen", fontface = "bold") +
  labs(title = paste0("100 training sets at x = ", round(x0, 1), " (Model A: degree 5)"),
       subtitle = "We re-train Model A on 100 different random samples. Each orange dot is one prediction at this x.",
       x = "x", y = "y", color = "Fit", shape = "Data") +
  theme(legend.position = "top")
```

--

**Variance** = spread of the orange dots around their mean. **Bias** = gap between the mean (X) and the true f(x).

---

# Model B (degree 8): Less Bias, More Variance

```{r validation-zoom-B, echo=FALSE, fig.height=4.2}
# sim_preds_B, ef0_B, sd0_B already computed in the Model A chunk
pred_B0 <- predict(fit_B, tibble(x = x0))

sim_df_B <- tibble(x = x0 + runif(n_sim, -0.15, 0.15), y = sim_preds_B)

p_simple_B <- ggplot() +
  geom_point(data = demo_train, aes(x, y, shape = "Training"), alpha = 0.3, size = 1.5) +
  geom_point(data = demo_val, aes(x, y, shape = "Validation"), color = "darkgreen", size = 2) +
  scale_shape_manual(values = c("Training" = 16, "Validation" = 17)) +
  geom_line(data = x_grid, aes(x, pred_B, color = "B: degree 8"), linewidth = 1) +
  stat_function(fun = sin, aes(color = "True f(x) = sin(x)"),
                linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(values = model_colors)

p_simple_B +
  geom_point(data = sim_df_B, aes(x = x, y = y), color = "orange", alpha = 0.5, size = 2.5) +
  geom_point(aes(x = x0, y = pred_B0), size = 4, color = "red", shape = 16) +
  geom_point(aes(x = x0, y = ef0_B), size = 6, shape = 4, color = "orange", stroke = 2) +
  coord_cartesian(xlim = c(x0 - 1.2, x0 + 1.2),
                  ylim = shared_ylim) +
  # Bias bracket
  annotate("segment", x = x0 + 0.3, xend = x0 + 0.3, y = f0, yend = ef0_B,
           color = "firebrick", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.22, xend = x0 + 0.38, y = f0, yend = f0,
           color = "firebrick", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.22, xend = x0 + 0.38, y = ef0_B, yend = ef0_B,
           color = "firebrick", linewidth = 0.8) +
  # Epsilon bracket
  annotate("segment", x = x0 - 0.3, xend = x0 - 0.3, y = f0, yend = y0,
           color = "darkgreen", linewidth = 0.8) +
  annotate("segment", x = x0 - 0.38, xend = x0 - 0.22, y = f0, yend = f0,
           color = "darkgreen", linewidth = 0.8) +
  annotate("segment", x = x0 - 0.38, xend = x0 - 0.22, y = y0, yend = y0,
           color = "darkgreen", linewidth = 0.8) +
  # Variance bracket
  annotate("segment", x = x0 + 0.6, xend = x0 + 0.6,
           y = ef0_B - sd0_B, yend = ef0_B + sd0_B,
           color = "steelblue", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.52, xend = x0 + 0.68,
           y = ef0_B - sd0_B, yend = ef0_B - sd0_B,
           color = "steelblue", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.52, xend = x0 + 0.68,
           y = ef0_B + sd0_B, yend = ef0_B + sd0_B,
           color = "steelblue", linewidth = 0.8) +
  # Color-coded legend in bottom-right
  annotate("rect", xmin = x0 + 0.72, xmax = x0 + 1.18,
           ymin = shared_ylim[1] + 0.01, ymax = shared_ylim[1] + 0.40,
           fill = "white", color = "grey40", linewidth = 0.4) +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.33, hjust = 0,
           label = paste0("E[f-hat(x)] = ", round(ef0_B, 2)),
           size = 3.5, color = "orange", fontface = "bold") +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.25, hjust = 0,
           label = paste0("Bias = ", round(ef0_B - f0, 2)),
           size = 3.5, color = "firebrick", fontface = "bold") +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.17, hjust = 0,
           label = paste0("SD = ", round(sd0_B, 3)),
           size = 3.5, color = "steelblue", fontface = "bold") +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.09, hjust = 0,
           label = paste0("epsilon = ", round(y0 - f0, 2)),
           size = 3.5, color = "darkgreen", fontface = "bold") +
  labs(title = paste0("100 training sets at x = ", round(x0, 1), " (Model B: degree 8)"),
       subtitle = "Compare with previous slide: bias shrinks, but the orange dots spread out (more variance).",
       x = "x", y = "y", color = "Fit", shape = "Data") +
  theme(legend.position = "top")
```

--

Bias is smaller, but variance is larger — the **tradeoff** in action.

---

# Model C (degree 20): Less Bias, More Variance

```{r validation-zoom-C, echo=FALSE, fig.height=4.2}
# sim_preds_C, ef0_C, sd0_C already computed in the Model A chunk
pred_C0 <- predict(fit_C, tibble(x = x0))

sim_df_C <- tibble(x = x0 + runif(n_sim, -0.15, 0.15), y = sim_preds_C)

p_simple_C <- ggplot() +
  geom_point(data = demo_train, aes(x, y, shape = "Training"), alpha = 0.3, size = 1.5) +
  geom_point(data = demo_val, aes(x, y, shape = "Validation"), color = "darkgreen", size = 2) +
  scale_shape_manual(values = c("Training" = 16, "Validation" = 17)) +
  geom_line(data = x_grid, aes(x, pred_C, color = "C: degree 20"), linewidth = 1) +
  stat_function(fun = sin, aes(color = "True f(x) = sin(x)"),
                linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(values = model_colors)

p_simple_C +
  geom_point(data = sim_df_C, aes(x = x, y = y), color = "orange", alpha = 0.5, size = 2.5) +
  geom_point(aes(x = x0, y = pred_C0), size = 4, color = "red", shape = 16) +
  geom_point(aes(x = x0, y = ef0_C), size = 6, shape = 4, color = "orange", stroke = 2) +
  coord_cartesian(xlim = c(x0 - 1.2, x0 + 1.2),
                  ylim = shared_ylim) +
  # Bias bracket
  annotate("segment", x = x0 + 0.3, xend = x0 + 0.3, y = f0, yend = ef0_C,
           color = "firebrick", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.22, xend = x0 + 0.38, y = f0, yend = f0,
           color = "firebrick", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.22, xend = x0 + 0.38, y = ef0_C, yend = ef0_C,
           color = "firebrick", linewidth = 0.8) +
  # Epsilon bracket
  annotate("segment", x = x0 - 0.3, xend = x0 - 0.3, y = f0, yend = y0,
           color = "darkgreen", linewidth = 0.8) +
  annotate("segment", x = x0 - 0.38, xend = x0 - 0.22, y = f0, yend = f0,
           color = "darkgreen", linewidth = 0.8) +
  annotate("segment", x = x0 - 0.38, xend = x0 - 0.22, y = y0, yend = y0,
           color = "darkgreen", linewidth = 0.8) +
  # Variance bracket
  annotate("segment", x = x0 + 0.6, xend = x0 + 0.6,
           y = ef0_C - sd0_C, yend = ef0_C + sd0_C,
           color = "steelblue", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.52, xend = x0 + 0.68,
           y = ef0_C - sd0_C, yend = ef0_C - sd0_C,
           color = "steelblue", linewidth = 0.8) +
  annotate("segment", x = x0 + 0.52, xend = x0 + 0.68,
           y = ef0_C + sd0_C, yend = ef0_C + sd0_C,
           color = "steelblue", linewidth = 0.8) +
  # Color-coded legend in bottom-right
  annotate("rect", xmin = x0 + 0.72, xmax = x0 + 1.18,
           ymin = shared_ylim[1] + 0.01, ymax = shared_ylim[1] + 0.40,
           fill = "white", color = "grey40", linewidth = 0.4) +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.33, hjust = 0,
           label = paste0("E[f-hat(x)] = ", round(ef0_C, 2)),
           size = 3.5, color = "orange", fontface = "bold") +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.25, hjust = 0,
           label = paste0("Bias = ", round(ef0_C - f0, 2)),
           size = 3.5, color = "firebrick", fontface = "bold") +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.17, hjust = 0,
           label = paste0("SD = ", round(sd0_C, 3)),
           size = 3.5, color = "steelblue", fontface = "bold") +
  annotate("text", x = x0 + 0.75, y = shared_ylim[1] + 0.09, hjust = 0,
           label = paste0("epsilon = ", round(y0 - f0, 2)),
           size = 3.5, color = "darkgreen", fontface = "bold") +
  labs(title = paste0("100 training sets at x = ", round(x0, 1), " (Model C: degree 20)"),
       subtitle = "Bias nearly gone, but orange dots spread widely — variance dominates.",
       x = "x", y = "y", color = "Fit", shape = "Data") +
  theme(legend.position = "top")
```

--

Most flexible model: bias vanishes but variance explodes — **overfitting**.

---

# Connecting the Pieces: MSE = Bias² + Variance + σ²

```{r mse-decomposition, echo=FALSE}
# Reuse the Monte Carlo results from the validation-results chunk
sigma2 <- 0.09  # we set noise = rnorm(n, 0, 0.3), so σ² = 0.3²

# Compute average MSE across all 500 simulated training sets (on interior points)
n_mc <- nrow(preds_A)
avg_mse_A <- mean(rowMeans((preds_A - matrix(f_true, n_mc, length(f_true), byrow = TRUE))^2))
avg_mse_B <- mean(rowMeans((preds_B - matrix(f_true, n_mc, length(f_true), byrow = TRUE))^2))
avg_mse_C <- mean(rowMeans((preds_C - matrix(f_true, n_mc, length(f_true), byrow = TRUE))^2))

tibble(
  Model = c("A (degree 5)", "B (degree 8)", "C (degree 20)"),
  `Bias²` = round(c(bv_A[1], bv_B[1], bv_C[1]), 4),
  Variance = round(c(bv_A[2], bv_B[2], bv_C[2]), 4),
  `σ²` = sigma2,
  `Bias²+Var+σ²` = round(c(bv_A[1]+bv_A[2], bv_B[1]+bv_B[2], bv_C[1]+bv_C[2]) + sigma2, 4),
  `Avg MSE (500 sims)` = round(c(avg_mse_A, avg_mse_B, avg_mse_C) + sigma2, 4),
  `Val MSE (1 sample)` = round(c(mean((demo_val$y[interior] - predict(fit_A, demo_val[interior,]))^2),
                       mean((demo_val$y[interior] - predict(fit_B, demo_val[interior,]))^2),
                       mean((demo_val$y[interior] - predict(fit_C, demo_val[interior,]))^2)), 4),
  `Train MSE` = round(c(mean((demo_train$y - predict(fit_A, demo_train))^2),
                         mean((demo_train$y - predict(fit_B, demo_train))^2),
                         mean((demo_train$y - predict(fit_C, demo_train))^2)), 4)
) |> knitr::kable()
```

--

- **Bias² + Var + σ²** matches **Avg MSE** (averaged over 500 simulations) — the decomposition works!

--

- **Val MSE (1 sample)** is what we observe in practice — a single draw, so it won't match exactly

--

- In practice, you **cannot** compute Bias² or Variance from one training set. The decomposition is a conceptual tool that explains *why* Val MSE has a U-shape

--

- σ² = 0.09 is the **floor** — no model can beat it (irreducible noise)

---
class: inverse, center, middle

# Discrimination Angle
### Statistical Bias ≠ Social Bias

---

# Two Kinds of Bias

**Statistical bias:** model's average prediction is off — it systematically over- or under-estimates.

$$\text{Bias} = f(x_0) - \mathbb{E}[\hat{f}(x_0)]$$

Fix: use a more flexible model.

--

**Social bias (disparate impact):** model treats groups differently in ways that cause harm — even if statistically unbiased overall.

Fix: ...it's complicated. That's the rest of this course.

---

# A Concrete Example

Imagine a wait-time prediction model:

--

- Predicts average wait = 5 min across **all** neighborhoods ✅ (low bias!)

--

- But: **3 min** in wealthy/white neighborhoods, **8 min** in low-income/Black neighborhoods

--

- The model isn't *wrong* — it accurately reflects the *system*

--

- The system itself encodes discrimination, and the model faithfully learns it

--

> **A statistically good model can be a socially harmful model.**

---

# How This Plays Out at Uber/Lyft

1. **Dispatch optimization** minimizes total wait → sends more drivers where demand is dense → underserves sparse/low-income areas

--

2. **Surge pricing** responds to supply-demand imbalance → areas with fewer drivers get higher prices → low-income riders pay more

--

3. **ETA prediction** trained on historical data → if service was worse in certain neighborhoods, the model predicts longer waits → drivers avoid those areas → **self-fulfilling prophecy**

---

# What to Always Ask

When evaluating any model:

--

1. What is the loss **overall**?

--

2. What is the loss **per group**?

--

3. Are error rates **equal** across groups?

--

This is the foundation for fairness metrics in Modules 11–12.

---
class: inverse, center, middle

# Exercise
### Let's See It in the Data

---

# Part 1: Synthetic Ride-Sharing Data

```{r data-gen}
set.seed(42)
n <- 2000

neighborhoods <- tibble(
  neighborhood  = c("Downtown", "Midtown", "Eastside", "Southside"),
  base_wait     = c(3, 4, 7, 9),
  driver_supply = c(50, 35, 15, 10),
  pct_minority  = c(0.20, 0.35, 0.70, 0.85),
  median_income = c(95000, 72000, 38000, 29000)
)

rides <- tibble(
  ride_id = 1:n,
  neighborhood = sample(
    neighborhoods$neighborhood, n, replace = TRUE,
    prob = c(0.35, 0.30, 0.20, 0.15)
  )
) |>
  left_join(neighborhoods, by = "neighborhood") |>
  mutate(
    hour = sample(0:23, n, replace = TRUE),
    is_peak = as.integer(hour %in% c(7, 8, 9, 17, 18, 19)),
    peak_penalty = is_peak * (10 - driver_supply / 10),
    wait_time = pmax(base_wait + peak_penalty + rnorm(n, 0, 1.5), 0.5)
  )
```

---

# Wait Time Distributions

```{r density-plot, echo=FALSE, fig.height=4}
ggplot(rides, aes(x = wait_time, fill = neighborhood)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Wait Time Distributions by Neighborhood",
    x = "Wait Time (minutes)", y = "Density"
  )
```

Downtown and Midtown cluster around 3–4 min. Eastside and Southside: 7–9 min.

---

# Summary by Neighborhood

```{r summary}
rides |>
  group_by(neighborhood) |>
  summarise(
    n_rides = n(), avg_wait = round(mean(wait_time), 1),
    pct_minority = first(pct_minority),
    median_income = scales::dollar(first(median_income)),
    .groups = "drop"
  ) |>
  arrange(avg_wait) |>
  knitr::kable()
```

---

# Part 2: Train / Val / Test Split

```{r split}
set.seed(123)
train_idx <- sample(1:n, 0.6 * n)
val_idx   <- sample(setdiff(1:n, train_idx), 0.2 * n)
test_idx  <- setdiff(1:n, c(train_idx, val_idx))

train <- rides[train_idx, ]
val   <- rides[val_idx, ]
test  <- rides[test_idx, ]
```

---

# Part 2: Fit Polynomials

Fit polynomials of degree 1 through 15 on `hour`:

```{r poly-fit}
degrees <- 1:15
results <- map_dfr(degrees, function(d) {
  fit <- lm(wait_time ~ poly(hour, d), data = train)
  tibble(
    degree    = d,
    train_mse = mean((train$wait_time - predict(fit, train))^2),
    val_mse   = mean((val$wait_time - predict(fit, val))^2)
  )
})
```

---

# The Classic Tradeoff Curve

```{r bv-plot, echo=FALSE, fig.height=4}
results |>
  pivot_longer(c(train_mse, val_mse), names_to = "set", values_to = "mse") |>
  mutate(set = ifelse(set == "train_mse", "Training", "Validation")) |>
  ggplot(aes(x = degree, y = mse, color = set)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  labs(
    title = "Bias-Variance Tradeoff: Polynomial Degree vs. MSE",
    subtitle = "Training error always decreases; validation error has a sweet spot",
    x = "Polynomial Degree", y = "Mean Squared Error", color = "Dataset"
  )
```

---

# Part 3: Disaggregate by Group

Fit the best model and check errors **per neighborhood**:

```{r disaggregate}
best_degree <- results$degree[which.min(results$val_mse)]
best_fit <- lm(wait_time ~ poly(hour, best_degree), data = train)

test |>
  mutate(predicted = predict(best_fit, test)) |>
  group_by(neighborhood) |>
  summarise(
    n = n(), avg_actual = round(mean(wait_time), 1),
    avg_predicted = round(mean(predicted), 1),
    mse = round(mean((wait_time - predicted)^2), 1),
    pct_minority = first(pct_minority), .groups = "drop"
  ) |>
  arrange(pct_minority) |>
  knitr::kable()
```

---

# Error Concentrates in Minority Areas

```{r mse-bar, echo=FALSE, fig.height=4}
test |>
  mutate(predicted = predict(best_fit, test)) |>
  group_by(neighborhood) |>
  summarise(
    mse = mean((wait_time - predicted)^2),
    pct_minority = first(pct_minority), .groups = "drop"
  ) |>
  ggplot(aes(x = reorder(neighborhood, pct_minority), y = mse, fill = pct_minority)) +
  geom_col() +
  scale_fill_gradient(low = "steelblue", high = "firebrick", labels = scales::percent) +
  labs(
    title = "Model Error is NOT Equal Across Neighborhoods",
    subtitle = "Same model, same overall MSE — but error concentrates in minority areas",
    x = "Neighborhood (ordered by % minority)", y = "MSE", fill = "% Minority"
  )
```

---

# Part 4: Add Neighborhood as a Feature

.small[
```{r better-model}
better_fit <- lm(wait_time ~ poly(hour, best_degree) + neighborhood, data = train)

test |>
  mutate(pred_v1 = predict(best_fit, test),
         pred_v2 = predict(better_fit, test)) |>
  group_by(neighborhood) |>
  summarise(
    avg_actual = round(mean(wait_time), 1),
    avg_pred_v1 = round(mean(pred_v1), 1),
    avg_pred_v2 = round(mean(pred_v2), 1),
    mse_v1 = round(mean((wait_time - pred_v1)^2), 1),
    mse_v2 = round(mean((wait_time - pred_v2)^2), 1),
    pct_minority = first(pct_minority), .groups = "drop"
  ) |>
  arrange(pct_minority) |>
  knitr::kable()
```
]

Model v2 is **more accurate** — but it has **learned the disparity**.

---
class: inverse

# The Key Questions

<br>

### 1. Should a model predict what **IS** or what **SHOULD BE**?

--

<br>

### 2. If we use these predictions for driver dispatch, do we perpetuate the very inequality we measured?

--

<br>

### 3. Is a model that ignores neighborhood "fairer" even though it's less accurate?

---

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td>The Learning Problem <i>(just finished)</i></td><td>✓ done</td></tr>
<tr><td><b>2</b></td><td><b><a href="../module-02/slides.html">Linear Models →</a></b></td><td>next</td></tr>
<tr><td>3</td><td><a href="../module-03/slides.html">Model Evaluation &amp; Selection</a></td><td>upcoming</td></tr>
<tr><td>4</td><td>Tree-Based Methods</td><td>upcoming</td></tr>
<tr><td>6</td><td>Unsupervised Learning</td><td>upcoming</td></tr>
<tr><td>7</td><td>Neural Networks Foundations</td><td>upcoming</td></tr>
<tr><td>11</td><td><a href="../module-11/slides.html">Fairness Frameworks &amp; Metrics</a></td><td>✓ done</td></tr>
<tr><td>12</td><td>Auditing &amp; Interpretability</td><td>upcoming</td></tr>
</table>
