---
title: "Module 11: Fairness Frameworks & Metrics"
subtitle: "When You Can't Be Fair in More Than One Way"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.two-col { display: flex; align-items: flex-start; gap: 1em; }
.col-narrow { flex: 1; }
.col-wide   { flex: 2; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
theme_set(theme_minimal(base_size = 14))
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Learning Problem</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">Linear Models</a></td><td>✓ done</td></tr>
<tr><td>3</td><td><a href="../module-03/slides.html">Model Evaluation &amp; Selection</a></td><td>✓ done</td></tr>
<tr><td>4</td><td>Tree-Based Methods</td><td>upcoming</td></tr>
<tr><td>6</td><td>Unsupervised Learning</td><td>upcoming</td></tr>
<tr><td>7</td><td>Neural Networks Foundations</td><td>upcoming</td></tr>
<tr><td><b>11</b></td><td><b>Fairness Frameworks &amp; Metrics</b> <i>(you are here)</i></td><td>← current</td></tr>
<tr><td>12</td><td>Auditing &amp; Interpretability</td><td>upcoming</td></tr>
</table>

---

# From Module 3: We Saw the Gaps

Module 3's audit showed that the same model could have:

- Equal AUC across groups
- Different precision, recall, FPR per group

--

That tells us **the model isn't behaving the same for everyone**. But it doesn't tell us:

--

- Which gap matters?
- What does "fair" actually mean?
- Can we close all the gaps at once?

--

This module gives you the formal answers, and one **mathematical impossibility result** that says some of those answers exclude each other.

---

# The Setup

For a binary classifier:

- $Y \in \{0, 1\}$ — actual outcome (e.g. driver accepted)
- $\hat{Y} \in \{0, 1\}$ — model's prediction
- $A$ — protected attribute (e.g. minority status), used **only** for auditing

--

A "fairness criterion" is some equality between groups defined on $Y$, $\hat{Y}$, and $A$.

--

There are several. The famous ones turn out to be **mutually incompatible**.

---

# Three Fairness Criteria

| Criterion | Equality required |
|---|---|
| **Demographic parity** | $P(\hat Y = 1 \mid A = 0) = P(\hat Y = 1 \mid A = 1)$ |
| **Equalized odds** | $P(\hat Y = 1 \mid Y = y, A = 0) = P(\hat Y = 1 \mid Y = y, A = 1)$ for $y \in \{0, 1\}$ |
| **Predictive parity** | $P(Y = 1 \mid \hat Y = 1, A = 0) = P(Y = 1 \mid \hat Y = 1, A = 1)$ |

--

Each one says something different about what "fair" means. Let's read each.

---

# 1. Demographic Parity

$$P(\hat Y = 1 \mid A = 0) = P(\hat Y = 1 \mid A = 1)$$

**The model says "accept" at the same rate for both groups.**

--

- Cares only about the model's outputs, **ignores the actual labels**
- Makes sense when the positive decision is itself the resource (a job offer, a loan, an Uber dispatch)

--

**Failure mode:** if base rates genuinely differ, hitting equal positive rates forces the model to *miss real positives* in one group or *flag real negatives* in the other.

---

# 2. Equalized Odds

$$P(\hat Y = 1 \mid Y = 1, A = 0) = P(\hat Y = 1 \mid Y = 1, A = 1) \quad \text{(equal TPR)}$$
$$P(\hat Y = 1 \mid Y = 0, A = 0) = P(\hat Y = 1 \mid Y = 0, A = 1) \quad \text{(equal FPR)}$$

**Among the actually-positive, catch the same fraction in each group. Among the actually-negative, falsely flag at the same rate.**

--

- Allows different positive rates *only if the truth says so*
- Weaker version, **equal opportunity**, only requires equal TPR

--

**Makes sense when:** there's a real outcome you care about and missing it has direct human cost (medical, fraud, dispatch).

---

# 3. Predictive Parity

$$P(Y = 1 \mid \hat Y = 1, A = 0) = P(Y = 1 \mid \hat Y = 1, A = 1)$$

**A "positive" prediction means the same thing in each group** — equal precision.

--

- Stronger version: **calibration within groups**. At every score $s$, the actual positive rate is $s$ in both groups.
- This is what Module 3's calibration discussion sets up.

--

**Makes sense when:** the model's score is interpreted directly by downstream users — risk scores, lending probabilities, recidivism estimates.

---

# The Impossibility Theorem

> **Chouldechova (2017) / Kleinberg, Mullainathan & Raghavan (2017):** if base rates differ between groups, $P(Y = 1 \mid A = 0) \neq P(Y = 1 \mid A = 1)$, and the classifier is not perfect, then **at most one of {demographic parity, equalized odds, predictive parity}** can hold.

--

This is **not a quirk of any model**. It is a mathematical fact about the joint distribution of $Y$, $\hat Y$, $A$.

--

If you enforce two, the third **must** break.

--

**This is why fairness is a choice, not a calculation.**

---

# The COMPAS Argument, Revisited

Module 3 showed how COMPAS triggered a famous fairness debate:

--

- **ProPublica** said: COMPAS is unfair — it has a **higher FPR for Black defendants** (innocent people flagged at twice the rate). Equalized odds is broken.

--

- **Northpointe** said: COMPAS is fair — a score of 7 means the same recidivism risk for Black and white defendants. **Predictive parity holds**.

--

- **Both were right.** Base rates differ; the impossibility theorem says you cannot satisfy both at once. Choosing which to break is a value judgment, not a math error.

---

# Empirical Demonstration: Same Data, Sweep the Threshold

```{r impossibility-data, include=FALSE}
set.seed(2024); n <- 5000
neighborhoods <- tibble(
  neighborhood   = c("Downtown","Midtown","Eastside","Southside"),
  base_accept    = c(0.85, 0.78, 0.55, 0.40),
  pct_minority   = c(0.20, 0.35, 0.70, 0.85)
)
requests <- tibble(
  request_id = 1:n,
  neighborhood = sample(neighborhoods$neighborhood, n, TRUE,
                        prob = c(0.35, 0.30, 0.20, 0.15))
) |>
  left_join(neighborhoods, by = "neighborhood") |>
  mutate(
    hour = sample(0:23, n, replace = TRUE),
    is_night = as.integer(hour >= 22 | hour <= 5),
    trip_distance_mi = pmax(rlnorm(n, 1.2, 0.6), 0.3),
    rider_rating = pmin(pmax(rnorm(n, 4.7, 0.4), 1), 5),
    logit_p = qlogis(base_accept) - 0.6 * is_night -
              0.15 * abs(trip_distance_mi - 4) +
              0.5 * (rider_rating - 4.7),
    accepted = rbinom(n, 1, plogis(logit_p)),
    is_minority = rbinom(n, 1, pct_minority)
  )

set.seed(123)
train_idx <- sample(1:n, 0.7 * n)
train <- requests[train_idx, ]
test  <- requests[-train_idx, ]
fit <- glm(accepted ~ neighborhood + is_night + trip_distance_mi + rider_rating,
           data = train, family = binomial)
test$pred_prob <- predict(fit, test, type = "response")

group_metrics <- function(df, t) {
  df |>
    mutate(pred = as.integer(pred_prob >= t)) |>
    group_by(is_minority) |>
    summarise(
      pos_rate  = mean(pred),
      tpr       = sum(pred == 1 & accepted == 1) / sum(accepted == 1),
      fpr       = sum(pred == 1 & accepted == 0) / sum(accepted == 0),
      precision = sum(pred == 1 & accepted == 1) / max(sum(pred == 1), 1),
      .groups   = "drop"
    )
}
```

```{r impossibility-plot, echo=FALSE, fig.height=4}
thresholds <- seq(0.05, 0.95, by = 0.02)
sweep <- map_dfr(thresholds, function(t) {
  g <- group_metrics(test, t)
  tibble(
    threshold = t,
    `Demographic parity` = abs(diff(g$pos_rate)),
    `Equalized odds (TPR)` = abs(diff(g$tpr)),
    `Equalized odds (FPR)` = abs(diff(g$fpr)),
    `Predictive parity`   = abs(diff(g$precision))
  )
})

sweep |>
  pivot_longer(-threshold, names_to = "criterion", values_to = "gap") |>
  ggplot(aes(threshold, gap, color = criterion)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c(
    "Demographic parity"   = "firebrick",
    "Equalized odds (TPR)" = "darkgreen",
    "Equalized odds (FPR)" = "steelblue",
    "Predictive parity"    = "orange")) +
  labs(title = "No threshold zeros every gap (driver-acceptance model)",
       subtitle = "Each curve = |minority - non-minority|",
       x = "Decision threshold",
       y = "|gap| between groups",
       color = NULL)
```

---

# How Do You Enforce a Criterion?

Three injection points:

--

**Pre-processing** — massage the training data (reweight, resample, transform features) before training.
- Pros: model-agnostic. Cons: throws away signal; the model can relearn proxies.

--

**In-processing** — modify the training objective (fairness penalty, adversarial debiasing).
- Pros: principled. Cons: needs custom training code.

--

**Post-processing** — adjust the **decision rule** per group on the trained model's scores.
- Pros: simplest, model-agnostic. Cons: explicitly uses the protected attribute at decision time, often illegal.

---

# Post-Processing: Group-Specific Thresholds

The simplest fix: pick a different threshold for each group.

--

```{r postproc-demo, echo=FALSE}
target <- mean(test$pred_prob >= 0.5)
threshold_for_group <- function(g, target) {
  ts <- seq(0, 1, by = 0.001)
  ts[which.min(abs(sapply(ts, \(t) mean(g$pred_prob >= t)) - target))]
}
t_maj <- threshold_for_group(filter(test, is_minority == 0), target)
t_min <- threshold_for_group(filter(test, is_minority == 1), target)

test$pred_pp <- ifelse(test$is_minority == 1,
                       as.integer(test$pred_prob >= t_min),
                       as.integer(test$pred_prob >= t_maj))

before <- group_metrics(test, 0.5)
after  <- test |>
  group_by(is_minority) |>
  summarise(
    pos_rate  = mean(pred_pp),
    tpr       = sum(pred_pp == 1 & accepted == 1) / sum(accepted == 1),
    precision = sum(pred_pp == 1 & accepted == 1) / max(sum(pred_pp == 1), 1),
    .groups = "drop"
  )

tibble(
  Metric = c("Demographic parity gap", "TPR gap (eq. opp.)", "Precision gap"),
  Before = round(c(diff(before$pos_rate), diff(before$tpr), diff(before$precision)), 3),
  After  = round(c(diff(after$pos_rate),  diff(after$tpr),  diff(after$precision)), 3)
) |> knitr::kable()
```

--

We **closed the demographic parity gap** by using thresholds `r round(t_maj, 2)` (majority) and `r round(t_min, 2)` (minority). But the **TPR gap got worse** — exactly what the impossibility theorem predicts.

---

# Accuracy vs Fairness Frontier

```{r frontier-plot, echo=FALSE, fig.height=4}
t_min_grid <- seq(0.05, 0.95, by = 0.02)
frontier <- map_dfr(t_min_grid, function(tm) {
  pred <- ifelse(test$is_minority == 1,
                 as.integer(test$pred_prob >= tm),
                 as.integer(test$pred_prob >= 0.5))
  pos_rates <- tapply(pred, test$is_minority, mean)
  tibble(minority_threshold = tm,
         accuracy = mean(pred == test$accepted),
         dp_gap = abs(pos_rates[1] - pos_rates[2]))
})

ggplot(frontier, aes(dp_gap, accuracy)) +
  geom_line(linewidth = 1, color = "firebrick") +
  geom_point(size = 1.4, color = "firebrick") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  labs(title = "Accuracy vs Fairness frontier",
       subtitle = "Each point = one minority-group threshold; majority fixed at 0.5",
       x = "Demographic parity gap (|minority - majority pos_rate|)",
       y = "Accuracy")
```

Every fairness intervention sits **somewhere on this curve**. Math draws the curve; humans pick the point.

---
class: inverse

# The Key Questions

<br>

### 1. Which fairness criterion are you protecting? Why that one?

--

<br>

### 2. What does enforcing it cost — in accuracy, *and* in the other criteria?

--

<br>

### 3. Who gets to make that choice? (Spoiler: it's not the data scientist alone.)

---

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Learning Problem</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">Linear Models</a></td><td>✓ done</td></tr>
<tr><td>3</td><td><a href="../module-03/slides.html">Model Evaluation &amp; Selection</a></td><td>✓ done</td></tr>
<tr><td>4</td><td>Tree-Based Methods</td><td>upcoming</td></tr>
<tr><td>6</td><td>Unsupervised Learning</td><td>upcoming</td></tr>
<tr><td>7</td><td>Neural Networks Foundations</td><td>upcoming</td></tr>
<tr><td>11</td><td>Fairness Frameworks &amp; Metrics <i>(just finished)</i></td><td>✓ done</td></tr>
<tr><td><b>12</b></td><td><b>Auditing &amp; Interpretability</b></td><td>next</td></tr>
</table>

Say **"start module 12"** when ready.
