---
title: "Module 4: Tree-Based Methods"
subtitle: "Surge Pricing, Neighborhood Effects, and Algorithmic Redlining"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.tiny { font-size: 65%; }
.two-col { display: flex; align-items: flex-start; gap: 1em; }
.col-narrow { flex: 1; }
.col-wide   { flex: 2; }
.highlight-box {
  background: #fff3e0;
  border-left: 4px solid #e65100;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
library(tidymodels)
library(rpart)
library(rpart.plot)
library(ranger)
library(xgboost)
theme_set(theme_minimal(base_size = 14))
set.seed(42)
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Learning Problem</a></td><td>done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">Linear Models</a></td><td>done</td></tr>
<tr><td>3</td><td><a href="../module-03/slides.html">Model Evaluation & Selection</a></td><td>done</td></tr>
<tr><td><b>4</b></td><td><b>Tree-Based Methods</b> <i>(you are here)</i></td><td>current</td></tr>
<tr><td>5</td><td>Unsupervised Learning</td><td>upcoming</td></tr>
<tr><td>6</td><td>Neural Networks Foundations</td><td>upcoming</td></tr>
<tr><td>7</td><td><a href="../module-07/slides.html">Fairness Frameworks & Metrics</a></td><td>done</td></tr>
<tr><td>8</td><td>Auditing & Interpretability</td><td>upcoming</td></tr>
</table>

---

# The Setup: Surge Pricing

Ride-sharing platforms use **surge multipliers** to balance supply and demand. When demand is high in an area, prices go up.

--

The model takes features like:
- Time of day, day of week
- Pickup neighborhood / zip code
- Recent demand in the area
- Number of available drivers nearby

--

.highlight-box[
**The fairness question:** if the model learns that certain neighborhoods consistently have high demand and low supply, it charges more there. Those neighborhoods often correlate with race and income. Is the algorithm **redlining**?
]

---

# Simulated Surge Data (1/2)

.small[
```{r sim-data-1}
n <- 2000
surge <- tibble(
  hour         = sample(0:23, n, replace = TRUE),
  day_of_week  = sample(1:7, n, replace = TRUE),
  neighborhood = sample(c("Downtown", "Midtown", "Uptown",
                           "Southside", "Westend"), n, replace = TRUE),
  demand_ratio = rnorm(n, mean = case_when(
    neighborhood == "Downtown"  ~ 1.8,
    neighborhood == "Midtown"   ~ 1.4,
    neighborhood == "Southside" ~ 1.6,
    neighborhood == "Westend"   ~ 0.9,
    TRUE ~ 1.1), sd = 0.3),
  drivers_nearby = rpois(n, lambda = case_when(
    neighborhood == "Downtown"  ~ 12,
    neighborhood == "Southside" ~ 4,
    neighborhood == "Westend"   ~ 5,
    TRUE ~ 8)),
  is_weekend = day_of_week %in% c(6, 7),
  is_rush    = hour %in% c(7:9, 17:19)
)
```
]

---
count: false

# Simulated Surge Data (2/2)

.small[
```{r sim-data-2}
surge <- surge |>
  mutate(
    surge_mult = 1 + 0.4 * demand_ratio - 0.05 * drivers_nearby +
      0.3 * is_rush + 0.15 * is_weekend + rnorm(n, sd = 0.15),
    surge_mult = pmax(surge_mult, 1.0)  # floor at 1x
  )
glimpse(surge)
```
]

---

# Part 1: Decision Trees

---

# A Single Decision Tree

*A decision tree asks a sequence of yes/no questions to split the data.*

.pull-left[
.small[
```{r single-tree, echo=TRUE}
tree_fit <- rpart(
  surge_mult ~ hour + demand_ratio +
    drivers_nearby + neighborhood +
    is_weekend + is_rush,
  data = surge,
  control = rpart.control(maxdepth = 2))
```

- At the root, the tree tried all 6 features — `neighborhood != Southside` gave the lowest RSS
- Within each child, it tried all 6 again (features **can** be reused). `is_rush` won at depth 2
]
]

.pull-right[
```{r single-tree-plot, echo=FALSE, fig.height=4.5, fig.width=5}
rpart.plot(tree_fit, roundint = FALSE, digits = 3,
  split.fun = function(x, labs, digits, varlen, faclen) {
    gsub("neighborhood = Downtown,Midtown,Uptown,Westend",
         "neighborhood != Southside", labs)
  })
```
]

---

# How Does a Split Get Chosen?

For regression trees, at each node the algorithm tries every feature and every threshold, and picks the one that minimizes the **residual sum of squares** (RSS) in the two child nodes.

--

For a candidate split into left ( $L$ ) and right ( $R$ ):

$$\small \text{RSS} = \sum_{i \in L} (y_i - \bar{y}_L)^2 + \sum_{i \in R} (y_i - \bar{y}_R)^2$$

--

**Example:** 6 rides with surge: **1.1** (Downtown), **1.2** (Midtown), **1.4** (Uptown), **1.5** (Westend), **2.0** (Southside), **2.1** (Southside)

| | `!= Southside` | `!= Southside, Westend` |
|---|---|---|
| **Left** | 1.1, 1.2, 1.4, 1.5 | 1.1, 1.2, 1.4 |
| **Right** | 2.0, 2.1 | 1.5, 2.0, 2.1 |
| $\bar{y}_L$ , $\bar{y}_R$ | 1.30,  2.05 | 1.23,  1.87 |
| **RSS** | 0.07 + 0.005 = **0.075** | 0.035 + 0.282 = 0.317 |

The algorithm picks `!= Southside` — more homogeneous groups, lower RSS. Note: the tree is **greedy** — it picks the locally best split without looking ahead.

---

# For Classification: Gini Impurity

Our surge example predicts a number (regression → RSS). But when predicting a **class** (e.g., "high surge" vs "low surge"), the split criterion is **Gini impurity**. $p_k$ = fraction of class $k$ in the node:

$$G = 1 - \sum_k p_k^2$$

--

Gini measures how **mixed** a node is (lower = purer = better):

| Node: 10 rides | Composition | Gini |
|---|---|---|
| All high surge | 10 high, 0 low → $p_{\text{high}}=1$ | $1 - 1^2 = 0$ (pure) |
| Even mix | 5 high, 5 low → $p_{\text{high}}=0.5$ | $1 - 0.5^2 - 0.5^2 = 0.5$ (worst) |
| Mostly one class | 9 high, 1 low → $p_{\text{high}}=0.9$ | $1 - 0.9^2 - 0.1^2 = 0.18$ |

--

The tree picks the split that produces the largest **decrease in weighted Gini** — same logic as RSS, but for categories instead of numbers.

---
count: false

# For Classification: Gini — Worked Example

10 rides labeled "high" or "low" surge. Two candidate splits on `neighborhood`:

| Ride | Neighborhood | Class |
|---|---|---|
| 1–3 | Downtown | high |
| 4–5 | Midtown | low |
| 6–7 | Southside | high |
| 8–10 | Southside | low |

--

.small[
| | `!= Southside` (L: 5, R: 5) | `!= Downtown` (L: 7, R: 3) |
|---|---|---|
| **Left** | 3 high, 2 low → $G = 1 - 0.6^2 - 0.4^2 = 0.48$ | 2 high, 5 low → $G = 1 - 0.29^2 - 0.71^2 = 0.41$ |
| **Right** | 2 high, 3 low → $G = 1 - 0.4^2 - 0.6^2 = 0.48$ | 3 high, 0 low → $G = 0$ (pure!) |
| **Weighted** | $\frac{5}{10}(0.48) + \frac{5}{10}(0.48) = 0.48$ | $\frac{7}{10}(0.41) + \frac{3}{10}(0) = \textbf{0.29}$ |

The tree picks `!= Downtown` — it isolates a pure node (all high surge).
]

---

# Overfitting: The Core Problem

An unconstrained tree grows until every leaf is pure (or has one observation). It **memorizes** the training data.

```{r overfit-demo, fig.height=3.5}
deep_tree <- rpart(surge_mult ~ hour + demand_ratio + drivers_nearby +
                     neighborhood + is_weekend + is_rush,
                   data = surge, control = rpart.control(cp = 0, minsplit = 2))
cat("Leaves in unconstrained tree:", sum(deep_tree$frame$var == "<leaf>"))
```

--

**Controls for complexity:**
- `maxdepth` — how deep the tree can grow
- `min_n` (= `minsplit`) — minimum observations to attempt a split
- `cp` (complexity parameter) — prune splits that don't improve fit by at least this much

---

# The Bias-Variance View

.pull-left[
**Shallow tree (depth = 2)**
- High bias: misses real patterns
- Low variance: stable across samples

```{r shallow, fig.height=3, echo=FALSE}
rpart.plot(rpart(surge_mult ~ demand_ratio + drivers_nearby,
  data = surge, control = rpart.control(maxdepth = 2)),
  roundint = FALSE, digits = 2)
```
]

.pull-right[
**Deep tree (depth = 8)**
- Low bias: captures everything
- High variance: different data $\to$ very different tree

```{r deep, fig.height=3, echo=FALSE}
rpart.plot(rpart(surge_mult ~ demand_ratio + drivers_nearby,
  data = surge, control = rpart.control(maxdepth = 8, cp = 0)),
  roundint = FALSE, digits = 2)
```
]

---

# Part 2: Random Forests

---

# From One Tree to Many

A single tree is **unstable** — remove a few data points and you get a completely different tree. Random forests fix this by averaging many trees. Two key ideas: **bagging** and **feature randomization**.

--

.pull-left[
**Bagging** (Bootstrap AGGregating)
- Draw B bootstrap samples (with replacement, same size as original)
- Fit one tree to each
- Average the predictions
]

.pull-right[
.small[
**Example:** predict surge for a Southside rush-hour ride

| | Bootstrap sample | Best split | Pred. |
|---|---|---|---|
| Tree 1 | emphasizes Downtown | `demand > 1.5` | 1.50 |
| Tree 2 | more Southside rides | `nbhd != South.` | 1.72 |
| Tree 3 | more rush-hour rides | `is_rush = 1` | 1.61 |
| **Bagged** | | | **1.61** |

Pred. = predicted surge for one specific ride (the leaf mean it falls into). Each tree sees different data → different split → different prediction. Averaging smooths out the noise.
]
]

---

# Feature Randomization

Bootstrap diversifies the **data** each tree sees. Feature randomization diversifies the **splits** — at *every* node, a fresh random subset of `mtry` features is drawn.

```
Tree 1 (bootstrap sample #1):
  Root:  random 3 of 6 → {demand, hour, is_weekend}    → split: demand < 1.5
  Left:  random 3 of 6 → {neighborhood, drivers, rush} → split: is_rush = 1
  Right: random 3 of 6 → {demand, neighborhood, rush}  → split: nbhd != Southside

Tree 2 (bootstrap sample #2):
  Root:  random 3 of 6 → {drivers, neighborhood, rush} → split: nbhd != Southside
  Left:  random 3 of 6 → {demand, hour, drivers}       → split: drivers > 6
  Right: random 3 of 6 → {hour, is_weekend, rush}      → split: is_rush = 1

Tree 3 (bootstrap sample #3):
  Root:  random 3 of 6 → {demand, neighborhood, hour}  → split: demand < 1.3
  Left:  random 3 of 6 → {drivers, rush, is_weekend}   → split: is_weekend = 1
  Right: random 3 of 6 → {demand, drivers, rush}       → split: drivers > 8
```

--

Without feature randomization, every tree would split on `demand_ratio` first (the strongest predictor) and all trees would look nearly identical. Forcing a random subset at each node creates **decorrelated** trees.

---

name: decorrelation

# Why Decorrelation Matters

<a href="#variance-proof" style="position:absolute; bottom:12px; left:40px; font-size:11px; background:#e8eaf6; padding:2px 8px; border-radius:3px; z-index:100;">Proof</a>

.pull-left[
**Without feature randomization** (bagging only):

Every tree splits on `demand_ratio` first. The trees are highly correlated.

Averaging $B$ correlated predictions:

$$\text{Var}(\bar{X}) = \rho \sigma^2 + \frac{1-\rho}{B}\sigma^2$$

High $\rho$ $\to$ averaging barely helps.
]

--

.pull-right[
**With feature randomization** (`mtry` < $p$):

Some trees are forced to split on `drivers_nearby` or `hour` first. The trees are less correlated.

Low $\rho$ $\to$ averaging **dramatically** reduces variance.

This is the key insight of Breiman (2001).
]

---

# Random Forest in R

.pull-left[
.small[
```{r rf-fit}
rf_spec <- rand_forest(
  mtry = 2,
  min_n = 5,              # min obs per leaf
  trees = 500) |>
  set_engine("ranger",
    importance = "impurity") |>
  set_mode("regression")

rf_wf <- workflow() |>
  add_recipe(recipe(
    surge_mult ~ hour + demand_ratio +
      drivers_nearby + neighborhood +
      is_weekend + is_rush,
    data = surge)) |>
  add_model(rf_spec)

rf_fit <- fit(rf_wf, data = surge)
```

Unlike a single tree, RF wants **deep trees** — high variance per tree is fine because averaging cancels it out.
]
]

.pull-right[
.small[
```{r rf-output}
extract_fit_parsnip(rf_fit)$fit
```
]
]

---

# RF Hyperparameters

| Parameter | What it controls | Typical default | Effect of increasing |
|-----------|-----------------|----------------|---------------------|
| `num.trees` (B) | Number of trees | 500 | More stable; no overfitting risk |
| `mtry` | Features per split | $\sqrt{p}$ (classif.) or $p/3$ (regr.) | Stronger trees but more correlated |
| `min.node.size` | Minimum n per leaf | 1 (classif.) or 5 (regr.) | Simpler trees, less overfitting |

--

.highlight-box[
Random forests **don't overfit** as you add more trees — more trees just means more stable averaging. 500 is almost always enough. The main thing to tune is `mtry`.
]

*Why no overfitting? Each individual tree does overfit (deep, memorizes its bootstrap sample). But each tree overfits to **different noise** (different bootstrap sample + different feature subsets). When you average them, the noise cancels out — only the real signal survives. Adding tree 501 is just one more independent vote; it can't make the average worse.*

---

# Variable Importance (RF)

```{r rf-importance, echo=FALSE, fig.height=3}
rf_raw <- extract_fit_parsnip(rf_fit)$fit
r2 <- 1 - rf_raw$prediction.error / var(surge$surge_mult)

tibble(
  variable = names(rf_raw$variable.importance),
  importance = rf_raw$variable.importance
) |>
  mutate(variable = fct_reorder(variable, importance)) |>
  ggplot(aes(importance, variable)) +
  geom_col(fill = "steelblue") +
  labs(title = paste0("RF Variable Importance (OOB R² = ", round(r2, 2), ")"),
       x = "Total RSS reduction across all trees", y = NULL)
```

.small[
**How to read this:** the R² tells you the model explains a lot (or a little) of the total variance. The bars show how that predictive work is distributed across features. `drivers_nearby` and `demand_ratio` do most of the heavy lifting.

**Caution:** this plot only shows **relative** importance (which features matter more than others). Two models can have identical importance rankings but very different R² — one explains 90% of variance, the other 5%. Always check R² first to know if the total prediction is meaningful.
]

---

# Part 3: Gradient Boosting

---

name: boosting

# Boosting: Sequential Correction

<a href="#gradient-proof" style="position:absolute; bottom:12px; left:40px; font-size:11px; background:#e8eaf6; padding:2px 8px; border-radius:3px; z-index:100;">Gradient proof</a>

Random forests build trees **independently**. Boosting builds them **sequentially** — each tree fixes the errors of the ensemble so far.

--

**Algorithm:**
1. Start with a constant prediction: $\hat{f}_0 = \bar{y}$
2. For $m = 1, \ldots, M$:
   - Compute residuals: $r_i = y_i - \hat{f}_{m-1}(x_i)$
   - Fit a shallow tree $h_m$ to the residuals (typically depth 1–6; depth 1 = a single split, called a "stump")
   - Update: $\hat{f}_m = \hat{f}_{m-1} + \eta \cdot h_m$

--

$\eta$ is the **learning rate** — how much each new tree contributes. Smaller $\eta$ = more conservative, needs more trees, but generalizes better.

--

**Why "gradient"?** The residual $r_i = y_i - \hat{f}(x_i)$ is the negative gradient of squared-error loss: $-\partial L / \partial \hat{f} = y_i - \hat{f}$. So "fit to residuals" = "fit to the negative gradient." For other losses (e.g., log-loss for classification), the gradient differs from a simple residual, but the idea is the same: gradient descent in function space.

---
count: false

# Boosting: Step-by-Step Example

$h_m$ is a tree — it takes features in, outputs a number (the leaf mean). $\eta \cdot h_m$ just scales that number down.

.small[
Tracking **one ride** (Southside, rush hour, actual = **1.80**). $\bar{y} = 1.35$, $\eta = 0.3$.

| Step | Residual | Leaf this ride falls into | $h_m$ (leaf mean) | Update |
|---|---|---|---|---|
| Start | 0.45 | | | $\hat{f}_0 = 1.35$ |
| Tree 1 | 0.45 | 3 rush rides: residuals 0.45, 0.56, 0.37 | 0.46 | $1.35 + 0.3 \times 0.46 = \textbf{1.49}$ |
| Tree 2 | 0.31 | 4 Southside rides: residuals 0.31, 0.28, 0.40, 0.37 | 0.34 | $1.49 + 0.3 \times 0.34 = \textbf{1.59}$ |
| Tree 3 | 0.21 | 3 high-demand rides: residuals 0.21, 0.15, 0.27 | 0.21 | $1.59 + 0.3 \times 0.21 = \textbf{1.65}$ |

$h_m$ is not this ride's residual — it's the **average residual in the leaf** (which contains other rides too). Only $\eta = 30\%$ of each correction is applied. The ensemble slowly converges: 1.35 → 1.49 → 1.59 → 1.65 → ... → 1.80.
]

---

# Boosting Toy Example (Visual)

```{r boost-toy, echo=FALSE, fig.height=4.2}
toy <- tibble(x = seq(0, 6, length.out = 50),
              y = sin(x) + rnorm(50, sd = 0.3))

boost_steps <- list()
pred <- rep(mean(toy$y), 50)

for (i in 1:4) {
  resid <- toy$y - pred
  fit_i <- rpart(resid ~ x, data = tibble(x = toy$x, resid = resid),
                 control = rpart.control(maxdepth = 1))
  correction <- predict(fit_i)
  pred <- pred + 0.3 * correction
  boost_steps[[i]] <- tibble(x = toy$x, ensemble = pred,
                              step = paste("After tree", i))
}

bind_rows(boost_steps) |>
  ggplot(aes(x, ensemble)) +
  geom_point(data = toy, aes(x, y), alpha = 0.3, size = 1) +
  geom_line(color = "firebrick", linewidth = 1) +
  facet_wrap(~step, nrow = 1) +
  labs(title = "Boosting: each tree corrects the previous ensemble",
       subtitle = "learning rate = 0.3, depth = 1 (stumps)",
       y = "Prediction")
```

---

# GBM Hyperparameters

| Parameter | What it controls | Typical range | Interaction |
|-----------|-----------------|--------------|------------|
| `trees` (M) | Boosting rounds | 100–1000 | More trees + low $\eta$ = better but slower |
| `tree_depth` | Complexity per tree | 1–6 | Depth 1 = main effects only; 6 = complex interactions |
| `learn_rate` ( $\eta$ ) | Shrinkage | 0.01–0.3 | Lower = needs more trees but generalizes better |

--

.highlight-box[
**Unlike RF, boosting CAN overfit** with too many trees. That's why we tune `trees` via cross-validation — the optimal number depends on `tree_depth` and `learn_rate`.
]

---

# XGBoost in R

.pull-left[
.small[
```{r xgb-fit}
gbm_spec <- boost_tree(
  trees      = 200,
  tree_depth = 3,
  learn_rate = 0.1) |>
  set_engine("xgboost",
    verbosity = 0) |>
  set_mode("regression")

gbm_wf <- workflow() |>
  add_recipe(recipe(
    surge_mult ~ hour + demand_ratio +
      drivers_nearby + neighborhood +
      is_weekend + is_rush,
    data = surge) |>
    step_mutate(across(where(is.logical), as.numeric)) |>
    step_dummy(all_nominal_predictors())) |>
  add_model(gbm_spec)

gbm_fit <- fit(gbm_wf, data = surge)
```
]
]

.pull-right[
.small[
- `step_dummy` converts `neighborhood` to 0/1 columns (XGBoost needs numeric input)
- Same `workflow()` pattern as RF — recipe handles preprocessing
- In practice, replace fixed values with `tune()` and use `tune_grid()` to find the best combination via CV
]
]

---

# XGBoost Variable Importance

```{r xgb-importance, echo=FALSE, fig.height=3.5}
xgb_imp <- xgb.importance(model = extract_fit_parsnip(gbm_fit)$fit)
xgb_imp |>
  mutate(Feature = fct_reorder(Feature, Gain)) |>
  ggplot(aes(Gain, Feature)) +
  geom_col(fill = "steelblue") +
  labs(title = "XGBoost Variable Importance (Gain)",
       x = "Fractional Gain", y = NULL)
```

---

# Part 4: Comparing the Three

---

# Predictions Across Neighborhoods

*Points on the dashed line = perfect prediction. Tighter cloud = better model. Color spread = neighborhood matters.*

```{r compare-preds, fig.height=3.5, echo=FALSE}
surge_test <- surge |> slice_sample(n = 500)

preds_compare <- surge_test |>
  mutate(
    Single_Tree = predict(tree_fit, newdata = surge_test),
    Random_Forest = predict(rf_fit, new_data = surge_test)$.pred,
    XGBoost = predict(gbm_fit, new_data = surge_test)$.pred
  ) |>
  pivot_longer(c(Single_Tree, Random_Forest, XGBoost),
               names_to = "model", values_to = "predicted") |>
  mutate(model = factor(model, levels = c("Single_Tree", "Random_Forest", "XGBoost")))

preds_compare |>
  ggplot(aes(surge_mult, predicted, color = neighborhood)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~model) +
  labs(title = "Actual vs Predicted Surge Multiplier",
       x = "Actual", y = "Predicted", color = "Neighborhood")
```

Single tree: blocky predictions (few distinct values). RF and XGBoost: smoother, tighter around the diagonal. But all three show **neighborhood-colored clusters** — the model treats neighborhoods differently.

---

# The Fairness Lens

```{r fairness-lens, fig.height=3.5, echo=FALSE}
preds_compare |>
  filter(model == "XGBoost") |>
  ggplot(aes(neighborhood, predicted, fill = neighborhood)) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "XGBoost Predicted Surge by Neighborhood",
       subtitle = "Southside and Downtown get systematically higher surge multipliers",
       x = NULL, y = "Predicted Surge Multiplier")
```

--

The model learned that **Southside** has high demand and few drivers $\to$ high surge. Is this efficient pricing or algorithmic redlining? That depends on *why* there are few drivers — which is Module 7's question.

---

# Part 5: SHAP Values

---

# Beyond Importance: Why Did *This* Ride Get a High Price?

Variable importance tells you which features matter **globally**. SHAP values tell you why the model made **this specific prediction**.

--

For each prediction:

$$\hat{f}(x_i) = \phi_0 + \phi_{\text{demand}}(x_i) + \phi_{\text{drivers}}(x_i) + \phi_{\text{neighborhood}}(x_i) + \ldots$$

--

- $\phi_0$ = average prediction across all rides
- $\phi_j(x_i)$ = how much feature $j$ pushed *this* ride's prediction above or below average
- All SHAP values sum exactly to the prediction (the **efficiency** property)

---

# SHAP for Our Surge Model

```{r shap-compute, fig.height=4}
# Extract raw xgboost object and prepare data matrix
xgb_raw <- extract_fit_parsnip(gbm_fit)$fit
surge_baked <- bake(extract_recipe(gbm_fit, estimated = TRUE),
                     new_data = surge) |>
  select(-surge_mult) |>
  mutate(across(where(is.logical), as.numeric)) |>
  as.matrix()
dtrain_shap <- xgb.DMatrix(surge_baked)

shap_contrib <- predict(xgb_raw, newdata = dtrain_shap, predcontrib = TRUE)
# Last column is BIAS (intercept) — drop it to keep only feature contributions
shap_mat <- shap_contrib[, -ncol(shap_contrib), drop = FALSE]
shap_df <- as_tibble(shap_mat) |>
  set_names(colnames(surge_baked))

# Mean absolute SHAP per feature
shap_importance <- shap_df |>
  summarise(across(everything(), ~ mean(abs(.)))) |>
  pivot_longer(everything(), names_to = "feature", values_to = "mean_abs_shap") |>
  mutate(feature = fct_reorder(feature, mean_abs_shap))

shap_importance |>
  slice_max(mean_abs_shap, n = 10) |>
  ggplot(aes(mean_abs_shap, feature)) +
  geom_col(fill = "steelblue") +
  labs(title = "Mean |SHAP| — Which Features Drive Predictions Most?",
       x = "Mean |SHAP value|", y = NULL)
```

---

# SHAP for Individual Rides

.small[
```{r shap-individual, fig.height=3.5}
# Pick a high-surge ride and a low-surge ride
high_idx <- which.max(surge$surge_mult)
low_idx  <- which.min(surge$surge_mult)

bind_rows(
  tibble(feature = colnames(surge_baked),
         shap = as.numeric(shap_mat[high_idx, ]),
         ride = paste("High surge:", round(surge$surge_mult[high_idx], 2))),
  tibble(feature = colnames(surge_baked),
         shap = as.numeric(shap_mat[low_idx, ]),
         ride = paste("Low surge:", round(surge$surge_mult[low_idx], 2)))
) |>
  group_by(ride) |>
  slice_max(abs(shap), n = 6) |>
  mutate(feature = fct_reorder(feature, abs(shap))) |>
  ggplot(aes(shap, feature, fill = shap > 0)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ride, scales = "free_y") +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "steelblue")) +
  labs(title = "SHAP Decomposition: Why Did This Ride Get This Price?",
       x = "SHAP value (contribution to prediction)", y = NULL)
```
]

---

# The Audit Question

SHAP lets you answer: **"Is neighborhood driving the prediction, after controlling for demand and supply?"**

--

If `neighborhoodSouthside` has a large positive SHAP value even for rides where `demand_ratio` and `drivers_nearby` are average, that's evidence the model is using neighborhood as more than a proxy for supply/demand.

--

.highlight-box[
This is the bridge to Module 7 (fairness) and Module 8 (auditing): tree-based models are powerful predictors but can encode geographic discrimination. SHAP is how you detect it.
]

---

# Key Takeaways

1. **Decision trees** are interpretable but overfit. Control with depth and pruning.

2. **Random forests** average many decorrelated trees $\to$ low variance without much bias increase. Main hyperparameter: `mtry`.

3. **Gradient boosting** (XGBoost) builds trees sequentially, correcting errors. Strongest off-the-shelf classifier, but **can overfit** — tune `trees`, `tree_depth`, `learn_rate` via CV.

4. **Variable importance** tells you what matters globally. **SHAP** tells you why each individual prediction was made.

5. All tree-based methods naturally handle **interactions** and **non-linearities** — no feature engineering needed.

---

# Exercise Preview

In the exercise you will:

1. Simulate a richer surge pricing dataset with explicit demographic correlations
2. Fit decision tree, random forest, and XGBoost models
3. Tune RF and XGBoost hyperparameters with cross-validation using `tidymodels`
4. Compare performance (RMSE, MAE) across models and neighborhoods
5. Compute SHAP values and check whether neighborhood effects persist after controlling for supply/demand
6. Write a 3-sentence "audit summary" of whether the model is redlining

See `exercise.R` for the starter code.

---
class: center, middle, inverse

# Backup Slides

---
name: variance-proof

# Backup: Variance of Averaged Correlated Predictions

<a href="#decorrelation" style="position:absolute; bottom:12px; left:40px; font-size:11px; background:#e8eaf6; padding:2px 8px; border-radius:3px; z-index:100;">Back</a>

We have $B$ trees, each producing prediction $X_i$ with $\text{Var}(X_i) = \sigma^2$ and pairwise correlation $\text{Cor}(X_i, X_j) = \rho$ for all $i \neq j$.

The ensemble prediction is $\bar{X} = \frac{1}{B}\sum_{i=1}^B X_i$. Its variance:

$$\text{Var}(\bar{X}) = \text{Var}\!\left(\frac{1}{B}\sum_i X_i\right) = \frac{1}{B^2} \sum_i \sum_j \text{Cov}(X_i, X_j)$$

--

Split the double sum into diagonal ( $i = j$ ) and off-diagonal ( $i \neq j$ ) terms:

$$= \frac{1}{B^2}\left[\sum_i \sigma^2 + \sum_{i \neq j} \rho\sigma^2\right] = \frac{1}{B^2}\left[B\sigma^2 + B(B-1)\rho\sigma^2\right]$$

--

$$= \frac{\sigma^2}{B} + \frac{B-1}{B}\rho\sigma^2 = \underbrace{\rho\sigma^2}\_{\text{can't reduce}} + \underbrace{\frac{(1-\rho)}{B}\sigma^2}\_{\to 0 \text{ as } B \to \infty}$$

As $B \to \infty$, only $\rho\sigma^2$ remains. **Lower $\rho$ (more decorrelated trees) = lower irreducible variance.**

---
name: gradient-proof

# Backup: Why "Gradient" Boosting?

<a href="#boosting" style="position:absolute; bottom:12px; left:40px; font-size:11px; background:#e8eaf6; padding:2px 8px; border-radius:3px; z-index:100;">Back</a>

.pull-left[
**Proof:** for squared-error loss $L = \frac{1}{2}(y_i - \hat{f}(x_i))^2$:

$$\frac{\partial L}{\partial \hat{f}(x_i)} = -(y_i - \hat{f}(x_i)) = -r_i$$

The **negative gradient** = the residual. So "fit a tree to the residuals" is gradient descent in function space: each tree steps in the direction that most reduces the loss.

For **log-loss** (classification): the gradient is $p_i - y_i$ (predicted prob minus label), which is *not* a simple residual — but the algorithm is the same: compute gradient → fit tree → step.
]

.pull-right[
```{r gradient-viz, echo=FALSE, fig.height=4.5, fig.width=5}
# Show loss curve and gradient direction for one observation
f_hat <- seq(0, 3, length.out = 200)
y_true <- 1.8
loss <- 0.5 * (y_true - f_hat)^2

# Current prediction
f_current <- 1.2
loss_current <- 0.5 * (y_true - f_current)^2
gradient <- -(y_true - f_current)  # negative, so step right

tibble(f_hat, loss) |>
  ggplot(aes(f_hat, loss)) +
  geom_line(linewidth = 1) +
  geom_point(aes(x = f_current, y = loss_current),
             color = "firebrick", size = 4) +
  geom_segment(aes(x = f_current, xend = f_current + 0.5,
                    y = loss_current, yend = loss_current),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "steelblue", linewidth = 1.5) +
  annotate("text", x = f_current + 0.25, y = loss_current + 0.02,
           label = "negative gradient\n(= residual = 0.6)",
           color = "steelblue", size = 3.5, vjust = -0.5) +
  annotate("text", x = f_current, y = loss_current + 0.04,
           label = paste0("current: f = ", f_current),
           color = "firebrick", size = 3.5, hjust = 1.1) +
  geom_vline(xintercept = y_true, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = y_true + 0.05, y = 0.15,
           label = "true y = 1.8", size = 3.5, hjust = 0) +
  labs(title = "Gradient descent on the loss curve",
       subtitle = "The residual points toward the minimum",
       x = expression(hat(f)(x[i])), y = "Loss") +
  theme_minimal(base_size = 12)
```
]
